---
title: "2. The CEF and Best Linear Predictor"
subtitle: "Why regression approximates the CEF"
---

```{r}
#| label: setup
#| message: false
library(ggplot2)
options(digits = 3)
```

The conditional expectation function $m(X) = \mathbb{E}[Y|X]$ is the best predictor of $Y$ given $X$ — but we rarely know it. The best linear predictor (BLP) is the next best thing: a linear approximation that only requires knowing means, variances, and covariances. This chapter explores the gap between the two and what we gain (and lose) by going linear.

**Questions this chapter answers:**

1. What is the best predictor of $Y$ given $X$, and why is it the conditional expectation function?
2. How does the best linear predictor (BLP) approximate a nonlinear CEF?
3. When does the BLP equal the CEF, and why does joint normality guarantee linearity?
4. How does omitted variable bias arise, and why don't additional controls always help?

::: {.callout-important}
## Population Quantities vs. Sample Estimates
The CEF and BLP are **population quantities** defined through expectations — $m(X) = \mathbb{E}[Y|X]$ and $\beta = \text{Var}(X)^{-1}\text{Cov}(X,Y)$ — and describe properties of the joint distribution of $(X,Y)$, not of any dataset. Running `lm()` in R performs OLS on a sample and produces *estimates* of the BLP. Running `loess()` produces a nonparametric smooth of sample data — an *estimate* of the CEF, not the CEF itself. In this chapter we often know the true data-generating process and can write down the CEF analytically (e.g., $m(X) = X^2$). When we can, we plot the true population CEF directly rather than approximating it from data.
:::

## CEF error: mean independence is not independence

Define the CEF error as $e = Y - m(X)$. By construction, $\mathbb{E}[e|X] = 0$ — this is called mean independence. But mean independence is weaker than full independence: the *variance* of $e$ can still depend on $X$.

Here's the lecture's example: $X \sim \text{Uniform}(-1, 1)$ and $Y = X^2 + X\varepsilon$, where $\varepsilon \sim N(0,1)$ is independent of $X$.

::: {#def-cef}
## Conditional Expectation Function
The CEF is $m(X) = \mathbb{E}[Y|X]$, the function that gives the mean of $Y$ for each value of $X$. The CEF error $e = Y - m(X)$ satisfies mean independence: $\mathbb{E}[e|X] = 0$.
:::

```{r}
#| label: mean-independence
set.seed(307)
n <- 5000
x <- runif(n, -1, 1)
eps <- rnorm(n)
y <- x^2 + x * eps

# CEF is m(X) = X^2
m_x <- x^2
e <- y - m_x

# Mean independence: E[e|X] ≈ 0 in every bin
bins <- cut(x, breaks = 20)
bin_means <- tapply(e, bins, mean)
round(bin_means, 3)
```

The conditional means of $e$ are all near zero. But the conditional *variance* grows with $|X|$:

```{r}
#| label: heteroskedastic-errors
df <- data.frame(x = x, e = e)

ggplot(df, aes(x, e)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue", linewidth = 1) +
  labs(x = "X", y = "CEF error (e)",
       title = "Mean independence holds, but errors are heteroskedastic",
       subtitle = "Var(e|X) = X², so spread increases with |X|") +
  theme_minimal()
```

The smooth line hovers at zero (mean independence), but the spread fans out. This pattern — $\text{Var}(e|X) = X^2$ — is heteroskedasticity. The CEF decomposes $Y$ into signal and noise, but doesn't guarantee the noise is uniform.

## The CEF is the best predictor {#sec-cef-best}

Among *all* functions of $X$, the CEF minimizes mean squared prediction error. Let's demonstrate by comparing the CEF to some alternatives for the same DGP.

```{r}
#| label: cef-best-predictor
# True CEF: m(X) = X^2
# Competitor 1: the unconditional mean
# Competitor 2: a linear predictor
# Competitor 3: a cubic predictor

mse_cef <- mean((y - x^2)^2)
mse_mean <- mean((y - mean(y))^2)
mse_linear <- mean(residuals(lm(y ~ x))^2)
mse_cubic <- mean(residuals(lm(y ~ x + I(x^2) + I(x^3)))^2)

data.frame(
  predictor = c("Unconditional mean", "OLS (linear)", "OLS (cubic)", "True CEF (X²)"),
  MSE = round(c(mse_mean, mse_linear, mse_cubic, mse_cef), 4)
)
```

The CEF has the lowest MSE. The cubic comes close because $X^3$ has mean zero for symmetric $X$, so the cubic approximation is nearly $X^2$. The linear predictor does worse — it can't capture the curvature.

## CEF vs. BLP: when the line misses the curve

The BLP minimizes $\mathbb{E}[(Y - a - bX)^2]$ and yields $b = \text{Cov}(X,Y)/\text{Var}(X)$. When the CEF is nonlinear, the BLP draws the best straight line through a curved relationship.

::: {#def-blp}
## Best Linear Predictor
The BLP of $Y$ given $X$ is $\alpha + \beta'X$ where $\beta = \text{Var}(X)^{-1}\text{Cov}(X, Y)$. The BLP error satisfies $\mathbb{E}[Xe] = 0$ (uncorrelated with $X$) but not necessarily $\mathbb{E}[e|X] = 0$.
:::

```{r}
#| label: cef-vs-blp
#| fig-cap: "The BLP is the best line through a curved CEF"
# Same DGP: Y = X^2 + X*eps
df <- data.frame(x = x, y = y)

ggplot(df, aes(x, y)) +
  geom_point(alpha = 0.1, size = 0.5) +
  stat_function(fun = function(x) x^2, aes(color = "True CEF: X²"),
                linewidth = 1.2) +
  stat_function(fun = function(x) 1/3, aes(color = "True BLP: 1/3"),
                linewidth = 1.2) +
  scale_color_manual(values = c("True CEF: X²" = "tomato",
                                "True BLP: 1/3" = "steelblue"),
                     name = "") +
  labs(x = "X", y = "Y",
       title = "The BLP is the best line through a curved CEF") +
  theme_minimal()
```

The true BLP is a flat line at $1/3$ because $\text{Cov}(X,Y) = \mathbb{E}[X^3] = 0$ by symmetry of $X$, giving slope $\beta = 0$ and intercept $\alpha = \mathbb{E}[Y] = \mathbb{E}[X^2] = 1/3$. The BLP does the best any line can, but it cannot capture the curvature.

The BLP residual $e = Y - \alpha - \beta X$ satisfies $\mathbb{E}[Xe] = 0$ (uncorrelated with $X$), but *not* $\mathbb{E}[e|X] = 0$. The residual still contains nonlinear structure: the conditional mean of the residual traces out $X^2 - 1/3$ rather than zero:

```{r}
#| label: blp-residual-pattern
# True BLP: α = 1/3, β = 0, so residual is Y − 1/3
df$blp_resid <- y - 1/3

ggplot(df, aes(x, blp_resid)) +
  geom_point(alpha = 0.1, size = 0.5) +
  stat_function(fun = function(x) x^2 - 1/3, color = "tomato", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "X", y = "BLP residual (Y − 1/3)",
       title = "BLP residuals still contain nonlinear structure",
       subtitle = "True E[e|X] = X² − 1/3 ≠ 0, but E[Xe] = 0") +
  theme_minimal()
```

## When does BLP = CEF?

The BLP equals the CEF when the CEF is actually linear. An important sufficient condition: if $(X, Y)$ are jointly normal, the CEF is linear. Let's verify.

```{r}
#| label: blp-equals-cef
library(MASS)
set.seed(42)

# Jointly normal (X, Y)
Sigma <- matrix(c(1, 0.7, 0.7, 1), 2, 2)
joint <- mvrnorm(5000, mu = c(0, 0), Sigma = Sigma)
x_norm <- joint[, 1]
y_norm <- joint[, 2]

df_norm <- data.frame(x = x_norm, y = y_norm)

ggplot(df_norm, aes(x, y)) +
  geom_point(alpha = 0.1, size = 0.5) +
  stat_function(fun = function(x) 0.7 * x, aes(color = "True CEF: 0.7X"),
                linewidth = 1.2) +
  geom_smooth(method = "lm", se = FALSE, aes(color = "OLS fit"),
              linewidth = 1.2, linetype = "dashed") +
  scale_color_manual(values = c("True CEF: 0.7X" = "tomato",
                                "OLS fit" = "steelblue"),
                     name = "") +
  labs(title = "Jointly normal: the true CEF is linear and OLS recovers it",
       x = "X", y = "Y") +
  theme_minimal()
```

The OLS line (dashed) sits on top of the true CEF because the CEF is genuinely linear when the data are jointly normal. We can verify analytically: with $\mu = 0$ and $\Sigma_{XY}/\Sigma_{XX} = 0.7$, the true CEF is $\mathbb{E}[Y|X] = 0.7X$, which is also the BLP. The OLS fit *estimates* this population line from the sample and, with $n = 5{,}000$, comes very close — but the population CEF and BLP exist before any data are drawn.

::: {.callout-note}
## Joint Normality Implies Linear CEF
If $(X, Y)$ are jointly normal, the conditional expectation $\mathbb{E}[Y|X]$ is exactly linear in $X$, and the BLP equals the CEF. This is a sufficient but not necessary condition — other distributions can also produce linear CEFs.
:::

## The BLP in formulas and code

The population BLP coefficient is $\beta = \text{Var}(X)^{-1}\text{Cov}(X, Y)$. In a sample, OLS estimates this:

```{r}
#| label: blp-formulas
# True population BLP: alpha = 0, beta = Sigma_XY / Sigma_XX = 0.7
# Sample analog using cov/var
beta_formula <- cov(x_norm, y_norm) / var(x_norm)
alpha_formula <- mean(y_norm) - beta_formula * mean(x_norm)

# OLS gives the same sample answer
beta_ols <- coef(lm(y_norm ~ x_norm))

cbind(population = c(0, 0.7),
      sample_analog = c(alpha_formula, beta_formula),
      OLS = beta_ols)
```

## Dummy variables: the CEF for discrete X

When $X$ is a categorical variable, the CEF is just group means. Regression with dummies recovers these exactly.

```{r}
#| label: dummies
set.seed(2026)
n <- 300
party <- sample(c("Dem", "Rep", "Ind"), n, replace = TRUE,
                prob = c(0.4, 0.35, 0.25))
approval <- 50 + 10 * (party == "Dem") - 5 * (party == "Rep") + rnorm(n, sd = 8)

df_party <- data.frame(party = factor(party), approval = approval)

# Group means = the CEF
tapply(df_party$approval, df_party$party, mean)

# Regression with dummies (R does this automatically)
coef(lm(approval ~ party, data = df_party))
```

The intercept is the mean for the baseline group (alphabetically first: Dem), and the coefficients are differences from that baseline.

## Interactions: when the slope depends on a group

An interaction term $D_i \times x_i$ allows the effect of $x$ to differ by group. The full model is

$$y = \beta_0 + \beta_1 x + \beta_2 d + \beta_3 (d \times x) + e$$

with four distinct behaviors baked into the four coefficients:

- $\beta_0$ is the intercept for the $d=0$ group
- $\beta_1$ is the slope on $x$ for the $d=0$ group
- $\beta_2$ is how much the $d=1$ intercept shifts up or down
- $\beta_3$ is how much the $d=1$ slope shifts relative to the $d=0$ slope

So the $d=1$ group has intercept $\beta_0 + \beta_2$ and slope $\beta_1 + \beta_3$. The **marginal effect of $d$ depends on $x$**: $\partial y / \partial d = \beta_2 + \beta_3 x$. The **marginal effect of $x$ depends on $d$**: $\partial y / \partial x = \beta_1 + \beta_3 d$.

### A zoo of interaction patterns

Six parameter patterns, six pictures. For each panel, blue is the $d=0$ line, red is the $d=1$ line. The same equation $y = \beta_0 + \beta_1 x + \beta_2 d + \beta_3 (d\times x)$ produces all six.

```{r}
#| label: interaction-zoo
#| fig-width: 9
#| fig-height: 5.5
models <- data.frame(
  model = factor(c("A", "B", "C", "D", "E", "F"),
                 levels = c("A", "B", "C", "D", "E", "F")),
  b0 = c(10, 1,  5,  5,  10, 10),
  b1 = c(0,  2,  0,  1,  -2, -2),
  b2 = c(-5, 1,  5,  0,  -8, -5),
  b3 = c(0,  -1, -1, -1, 4,  2)
)

grid <- expand.grid(x = seq(0, 5, length.out = 2),
                    d = c(0, 1),
                    model = models$model)
grid <- merge(grid, models, by = "model")
grid$y <- with(grid, b0 + b1 * x + b2 * d + b3 * d * x)
grid$d <- factor(grid$d, labels = c("d = 0", "d = 1"))

ggplot(grid, aes(x, y, color = d)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ model, ncol = 3) +
  scale_color_manual(values = c("d = 0" = "steelblue",
                                "d = 1" = "tomato")) +
  coord_cartesian(ylim = c(0, 11)) +
  labs(x = "x", y = "y", color = NULL,
       title = "Six interaction patterns from the same equation")
```

Every one of these shapes — parallel lines, converging lines, crossing lines, one flat and one sloped — is a combination of $(\beta_0, \beta_1, \beta_2, \beta_3)$. The shape tells you which coefficients are zero and which are not:

|   | $\beta_1$ | $\beta_2$ | $\beta_3$ | Pattern |
|---|---|---|---|---|
| A | 0 | $\neq 0$ | 0 | Parallel, both flat, $d=1$ shifted down |
| B | $\neq 0$ | $\neq 0$ | $\neq 0$ | Both slope up, $d=1$ less steep, crossing |
| C | 0 | $\neq 0$ | $\neq 0$ | $d=0$ flat, $d=1$ starts high and decays |
| D | $\neq 0$ | 0 | $\neq 0$ | Same intercept, diverging slopes |
| E | $\neq 0$ | $\neq 0$ | $\neq 0$ | Opposite slopes: crossing, "no main effect" illusion |
| F | $\neq 0$ | $\neq 0$ | $\neq 0$ | $d=1$ flat while $d=0$ decreases |

Pattern E is the classic cautionary case: at $x \approx 2$ both lines cross, so the "average effect of $d$" is near zero. But $d$ matters greatly — its effect just depends on $x$. A $t$-test on $\beta_2$ alone would miss this; the joint test of $\beta_2 = 0, \beta_3 = 0$ does not. See the [training × experience example](ch06-small-sample.qmd#a-motivating-example-does-training-affect-wages) in Chapter 6.

### Mapping theories to parameter restrictions

Many substantive claims — from political science, policy analysis, and the social sciences broadly — are claims about *conjunctions*: "$x$ matters only when $d$ is present", "$d$ is necessary for anything to happen", "$x$ and $d$ substitute for each other." The interaction model turns each of these into a testable restriction on $(\beta_1, \beta_2, \beta_3)$.

| Theoretical claim | Restriction on $\beta$ | Type |
|---|---|---|
| $d$ has no effect at all (is irrelevant) | $\beta_2 = 0$ **and** $\beta_3 = 0$ | joint ($q=2$) |
| $x$ has no effect at all | $\beta_1 = 0$ **and** $\beta_3 = 0$ | joint ($q=2$) |
| $d$ is necessary for $x$ to matter | $\beta_1 = 0$, $\beta_3 \neq 0$ | single + sign |
| $x$ is necessary for $d$ to matter | $\beta_2 = 0$, $\beta_3 \neq 0$ | single + sign |
| Effects are additive (no interaction) | $\beta_3 = 0$ | single |
| $d$ is sufficient: $y$ saturates when $d = 1$ | $\beta_1 + \beta_3 = 0$ | linear combination |
| $x$ and $d$ are complements (reinforcing) | $\beta_3 > 0$ | sign |
| $x$ and $d$ are substitutes (diminishing) | $\beta_3 < 0$ | sign |

The first two rows are the crucial ones for practice. A theory that says "this variable is irrelevant" is **never** a single-coefficient claim once an interaction is in the model. Testing it requires a **joint** hypothesis on two coefficients — which is exactly what the $F$-test and the robust Wald test in [Chapter 6](ch06-small-sample.qmd#the-f-test-when-one-t-test-isnt-enough) and [Chapter 9](ch09-testing.qmd#testing-multiple-restrictions-the-f-test) provide.

### A quick fit on simulated data

```{r}
#| label: interactions
set.seed(2026)
n_int <- 400
group <- sample(0:1, n_int, replace = TRUE)
x_int <- rnorm(n_int)
# Truth: group 0 has slope 2; group 1 has slope -1; intercept shift 1.5
y_int <- 3 + 1.5 * group + 2 * x_int - 3 * group * x_int + rnorm(n_int)

df_int <- data.frame(x = x_int, y = y_int, group = factor(group))

mod_parallel <- lm(y ~ group + x, data = df_int)        # β₃ = 0 imposed
mod_interact <- lm(y ~ group * x, data = df_int)        # unrestricted

coef(mod_interact)
```

The `group1:x` coefficient ($\approx -3$) is $\hat\beta_3$, the difference in slopes between group 1 and group 0. The marginal effect of $x$ in group 1 is $\hat\beta_1 + \hat\beta_3 \approx -1$; in group 0 it is $\hat\beta_1 \approx 2$.

Testing "group doesn't matter at all" is a joint hypothesis on the group dummy *and* its interaction:

```{r}
#| label: interactions-joint-test
#| message: false
library(car)
linearHypothesis(mod_interact, c("group1 = 0", "group1:x = 0"))
```

The point is correspondence: the hypothesis you test has to match the theory you wrote down. A $t$-test on $\beta_2 = 0$ is a test of "$d$ has no effect *at $x = 0$*" --- a specific claim about one point in the covariate space, not about whether $d$ matters in general. The joint test of $\beta_2 = \beta_3 = 0$ is what corresponds to "$d$ is irrelevant across the board." Different theories pick out different restrictions:

- "$d$ matters only when $x$ is above some level $x^*$": $\beta_2 + \beta_3 x^* = 0$ — a single linear restriction, not two.
- "$d$ and $x$ have identical effects at every value": $\beta_1 = \beta_2$ *and* $\beta_3 = 0$ — two restrictions.
- "Adding $d$ provides no information beyond $x$": $\beta_2 = \beta_3 = 0$ — the row from the table above.

## Omitted variable bias {#sec-ovb}

If the true model is $Y = X_1'\beta_1 + X_2'\beta_2 + e$ but we only regress on $X_1$, we get $\gamma_1 = \beta_1 + \Gamma_{12}\beta_2$, where $\Gamma_{12}$ is the regression of $X_2$ on $X_1$. The bias has a clear direction when we know the signs.

```{r}
#| label: ovb-simulation
set.seed(2026)
n_ovb <- 1000

# X2 (institutions) is correlated with X1 (democracy)
x1 <- rnorm(n_ovb)                        # democracy
x2 <- 0.6 * x1 + rnorm(n_ovb, sd = 0.8)  # institutions (correlated)
y_ovb <- 1 + 0.5 * x1 + 2 * x2 + rnorm(n_ovb)  # growth

# True model
coef(lm(y_ovb ~ x1 + x2))

# Short regression (omit x2)
coef(lm(y_ovb ~ x1))
```

The true effect of democracy ($\beta_1$) is 0.5, but the short regression gives $\gamma_1 \approx 1.7$. The bias is $\Gamma_{12} \times \beta_2 \approx 0.6 \times 2 = 1.2$, inflating the estimate because institutions are positively correlated with democracy and positively affect growth.

```{r}
#| label: ovb-formula
# Verify the OVB formula
gamma_12 <- coef(lm(x2 ~ x1))[2]  # regression of X2 on X1
beta_2 <- coef(lm(y_ovb ~ x1 + x2))[3]

c(short_coef = coef(lm(y_ovb ~ x1))[2],
  long_coef_plus_bias = coef(lm(y_ovb ~ x1 + x2))[2] + gamma_12 * beta_2)
```

The OVB formula $\gamma_1 = \beta_1 + \Gamma_{12}\beta_2$ matches exactly.

::: {#thm-ovb}
## Omitted Variable Bias
If the true model is $Y = X_1'\beta_1 + X_2'\beta_2 + e$ and we regress $Y$ on $X_1$ alone, the short regression coefficient is $\gamma_1 = \beta_1 + \Gamma_{12}\beta_2$, where $\Gamma_{12}$ is the coefficient from regressing $X_2$ on $X_1$. The bias $\Gamma_{12}\beta_2$ depends on both the correlation between $X_1$ and $X_2$ and the effect of $X_2$ on $Y$.
:::

::: {.callout-warning}
## OVB Is About Population, Not Sample
The omitted variable bias formula $\gamma_1 = \beta_1 + \Gamma_{12}\beta_2$ is an exact population relationship between projection coefficients. It does not require large samples, normality, or any distributional assumption — it follows purely from the algebra of linear projection.
:::

### Adding variables doesn't always help

A subtlety from the lecture: if the true model has three variables and you can't observe $X_3$, adding $X_2$ can make the bias on $\beta_1$ worse, depending on the correlation structure.

```{r}
#| label: ovb-worse
set.seed(2026)
n_ow <- 1000
z1 <- rnorm(n_ow)
z2 <- 0.3 * z1 + rnorm(n_ow)
z3 <- -0.8 * z1 + 0.5 * z2 + rnorm(n_ow)
y_ow <- 1 + 1 * z1 + 0.5 * z2 + 3 * z3 + rnorm(n_ow)

# True beta_1 = 1
c(true = coef(lm(y_ow ~ z1 + z2 + z3))[2],
  short = coef(lm(y_ow ~ z1))[2],
  medium = coef(lm(y_ow ~ z1 + z2))[2])
```

Why does adding $X_2$ make things worse? Apply the OVB formula to the medium regression: omitting $X_3$ biases $\hat\beta_1$ by $\Gamma_{13\cdot 2}\beta_3$, where $\Gamma_{13\cdot 2}$ is the coefficient on $X_1$ in the regression of $X_3$ on $(X_1, X_2)$. Conditioning on $X_2$ can change the partial correlation between $X_1$ and $X_3$, amplifying the bias.

```{r}
#| label: ovb-worse-decomposition
# OVB decomposition for the medium model
gamma_13_2 <- coef(lm(z3 ~ z1 + z2))[2]  # partial effect of z1 on z3, controlling for z2
beta_3 <- coef(lm(y_ow ~ z1 + z2 + z3))[4]

c(medium_bias = coef(lm(y_ow ~ z1 + z2))[2] - 1,
  ovb_formula = gamma_13_2 * beta_3)
```

Depending on the correlation structure, the "medium" model (with $X_2$ but not $X_3$) may be further from the truth than the "short" model. More controls are not automatically better.

## Causal interpretation: regression as a weighted average

The lecture ends with a deep result: when you regress $Y$ on a treatment dummy $D$ and controls $X$, the regression coefficient is a weighted average of the group-specific treatment effects $\delta_X = \mathbb{E}[Y|X, D=1] - \mathbb{E}[Y|X, D=0]$, with weights proportional to the conditional variance of $D$ given $X$.

$$\delta_R = \frac{\mathbb{E}[\sigma^2_{D|X} \cdot \delta_X]}{\mathbb{E}[\sigma^2_{D|X}]}$$ {#eq-blp}

Groups where treatment is most variable (closest to a 50/50 split) get the most weight.

```{r}
#| label: causal-weights
set.seed(2026)
n_cw <- 5000

# X determines both treatment probability and treatment effect
x_cw <- runif(n_cw, 0, 1)
prob_treat <- x_cw^2  # treatment rare for low X, common for high X
D <- rbinom(n_cw, 1, prob_treat)

# Heterogeneous treatment effect: large for low X, negative for high X
delta_x <- 10 - 16 * x_cw   # ranges from +10 at X=0 to -6 at X=1
y_cw <- 2 + 3 * x_cw + delta_x * D + rnorm(n_cw)

# Regression coefficient
delta_R <- coef(lm(y_cw ~ D + x_cw))[2]

# Weighted average with weights = Var(D|X) = p(1-p)
wt <- prob_treat * (1 - prob_treat)
delta_weighted <- sum(wt * delta_x) / sum(wt)

c(regression = delta_R, weighted_avg = delta_weighted,
  unweighted_avg = mean(delta_x))
```

The unweighted average treatment effect is about 2, but the regression coefficient is near 0. That's because treatment assignment $p(X) = X^2$ is most variable around $X \approx 0.7$ (where $p(1-p)$ peaks), and at that point the treatment effect $\delta(X) = 10 - 16(0.7) \approx -1$ is near zero or negative. The regression upweights these observations and downweights the large positive effects at low $X$, where treatment is rare and $\text{Var}(D|X)$ is small.

```{r}
#| label: causal-weights-plot
df_cw <- data.frame(x = x_cw, delta = delta_x, weight = wt)

ggplot(df_cw, aes(x)) +
  geom_line(aes(y = delta, color = "Treatment effect δ(X)"), linewidth = 1.2) +
  geom_line(aes(y = weight * 40, color = "Weight: Var(D|X) (rescaled)"),
            linewidth = 1.2) +
  scale_y_continuous(
    name = "Treatment effect δ(X)",
    sec.axis = sec_axis(~ . / 40, name = "Var(D|X)")
  ) +
  scale_color_manual(values = c("Treatment effect δ(X)" = "tomato",
                                "Weight: Var(D|X) (rescaled)" = "steelblue"),
                     name = "") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(x = "X", title = "Regression weights favor where treatment is most variable") +
  theme_minimal()
```

This is why a randomized experiment with $p = 0.5$ for everyone gives equal weight to all subgroups: when $\text{Var}(D|X)$ is constant, the variance-weighted average reduces to the simple average.

## Summary

- The **CEF** $m(X) = \mathbb{E}[Y|X]$ is the best predictor, but the CEF error can be heteroskedastic (mean independence $\neq$ independence).
- The **BLP** $X'\beta$ is the best *linear* approximation. It equals the CEF when the CEF is linear (e.g., jointly normal data).
- BLP residuals satisfy $\mathbb{E}[Xe] = 0$ (weaker than $\mathbb{E}[e|X] = 0$), so nonlinear patterns can remain.
- **Dummy variables** make the CEF discrete: the regression recovers group means.
- **Interactions** allow slopes to vary by group or by the value of a moderator.
- **Omitted variable bias** has a precise formula: $\gamma_1 = \beta_1 + \Gamma_{12}\beta_2$. Adding controls doesn't always reduce bias.
- **Causal interpretation**: regression weights treatment effects by the conditional variance of treatment assignment.

Next: [Multivariate OLS](ch03-ols.qmd) — moving from population quantities to sample estimation.
