---
title: "3. Multivariate OLS"
subtitle: "Deriving and computing the OLS estimator"
---

```{r}
#| label: setup
#| message: false
library(ggplot2)
library(carData)
options(digits = 3)
tr <- function(M) sum(diag(M))  # R has no built-in trace
```

The OLS estimator $\hat\beta = (X'X)^{-1}X'y$ is a matrix formula. Every piece of it — the transpose, the product, the inverse — has a direct R function. This chapter builds OLS from those building blocks, first geometrically in two dimensions, then in full matrix form with real data.

**Questions this chapter answers:**

1. What R functions implement matrix operations, and how does `crossprod()` relate to the normal equations?
2. What is the geometry of OLS — why is regression a projection?
3. How do we derive $\hat\beta = (X'X)^{-1}X'y$ from the minimization of SSE?
4. How do standard errors arise from $s_e^2(X'X)^{-1}$, and what makes them precise or imprecise?

## R's matrix toolkit

Before deriving anything, here are the operations we'll use throughout.

```{r}
#| label: matrix-toolkit
A <- matrix(c(2, 1, 1, 3), nrow = 2)
A

# 1. Transpose: t()
t(A)

# 2. Matrix multiply: %*%  (not * which is element-wise)
B <- matrix(c(1, 0, -1, 2), nrow = 2)
A %*% B

# 3. Inverse: solve()
solve(A)
A %*% solve(A)  # identity

# 4. Trace: sum(diag())
tr(A)   # sum of diagonal elements

# 5. Eigendecomposition: eigen()
eigen(A)

# 6. Determinant: det()
det(A)

# 7. Cross product shortcuts
x_vec <- c(1, 2, 3)
y_vec <- c(4, 5, 6)
c(manual = sum(x_vec * y_vec),
  crossprod = as.numeric(crossprod(x_vec, y_vec)))
```

`crossprod(X)` computes $X'X$ and `crossprod(X, y)` computes $X'y$. We'll use these constantly — and they are faster than the explicit `t(X) %*% X`. Why?

**Why `crossprod` is faster.** The expression `t(X) %*% X` does two things: first it allocates a new $K \times n$ matrix for the transpose, then it calls a general matrix multiply (DGEMM in the BLAS library) that treats the two inputs as unrelated. `crossprod(X)` skips the transpose allocation entirely and calls a specialized routine (DSYRK) that exploits the fact that $X'X$ is symmetric — it only computes the upper triangle, roughly halving the floating-point work.

```{r}
#| label: crossprod-benchmark
# Benchmark on a moderately large matrix
n_bench <- 5000
p_bench <- 50
X_bench <- matrix(rnorm(n_bench * p_bench), n_bench, p_bench)

t_explicit <- system.time(for (i in 1:200) t(X_bench) %*% X_bench)[["elapsed"]]
t_crossprod <- system.time(for (i in 1:200) crossprod(X_bench))[["elapsed"]]

c(explicit_sec = t_explicit,
  crossprod_sec = t_crossprod,
  speedup = t_explicit / t_crossprod)
```

The speedup depends on the BLAS implementation your R uses (Apple Accelerate, OpenBLAS, MKL, etc.) and on the matrix dimensions. For our course-sized examples ($n$ in the hundreds) the difference is negligible, but it's good practice — and when $n$ reaches the tens of thousands it adds up.

## Geometry: projection in two dimensions

You're used to plotting data with variables on the axes — one axis for $X$, one for $Y$, and each point is an observation. The geometric view of regression flips this: each axis is an *observation*, and each variable is a *vector*. A variable with $n$ observations is a vector in $\mathbb{R}^n$.

Why think this way? Because regression asks: among all scalar multiples of $\mathbf{x}$ (all predictions of the form $b\mathbf{x}$), which one is closest to $\mathbf{y}$? That's a projection — dropping a perpendicular from $\mathbf{y}$ onto the line spanned by $\mathbf{x}$.

With just two observations, we can see this on a page. Suppose we survey two students: we record how long each spent on homework ($\mathbf{h}$) and how long reading the textbook ($\mathbf{t}$). Each variable is a 2-vector, and we can plot both in $\mathbb{R}^2$:

```{r}
#| label: two-vectors
# Student 1: 3 hrs homework, 2.5 hrs reading
# Student 2: 5 hrs homework, 2 hrs reading
h <- c(3, 5)    # outcome: homework time
tt <- c(2.5, 2)  # predictor: reading time
```

We want to predict $\mathbf{h}$ using $\mathbf{t}$: find $b$ so that $b\mathbf{t}$ is as close to $\mathbf{h}$ as possible. Geometrically, $b\mathbf{t}$ must lie on the line through $\mathbf{t}$ (the dotted line in the plot below), and the best choice is the one where the "miss" — the residual — is perpendicular to that line.

The **vector projection** of $\mathbf{h}$ onto $\mathbf{t}$ solves this:

$$\hat{\mathbf{h}} = \frac{\mathbf{t} \cdot \mathbf{h}}{\mathbf{t} \cdot \mathbf{t}} \mathbf{t}$$

The scalar $b = \frac{\mathbf{t} \cdot \mathbf{h}}{\mathbf{t} \cdot \mathbf{t}}$ minimizes $\|\mathbf{h} - b\mathbf{t}\|^2$ — the sum of squared errors. This is the least squares solution, derived purely from geometry.

```{r}
#| label: projection-by-hand
# The projection coefficient
b <- as.numeric(crossprod(tt, h) / crossprod(tt))
b

# The projected vector (fitted values)
h_hat <- b * tt
h_hat
```

And this is exactly what `lm()` computes when we regress $\mathbf{h}$ on $\mathbf{t}$ without an intercept:

```{r}
#| label: projection-equals-ols
# OLS without intercept gives the same b
coef(lm(h ~ tt - 1))

# Fitted values = the projection
cbind(projection = h_hat, fitted = fitted(lm(h ~ tt - 1)))
```

The residual $\mathbf{e} = \mathbf{h} - \hat{\mathbf{h}}$ is orthogonal to $\mathbf{t}$ — their dot product is zero:

```{r}
#| label: residual-orthogonal
e_vec <- h - h_hat
as.numeric(crossprod(tt, e_vec))
```

This orthogonality is the geometric content of the normal equations $X'e = 0$. Here's the full picture:

```{r}
#| label: geometry-plot
#| fig-cap: "OLS finds the closest point on the line of t to h"
#| fig-width: 5
#| fig-height: 5
df_arrows <- data.frame(
  x0 = c(0, 0, 0, h_hat[1]),
  y0 = c(0, 0, 0, h_hat[2]),
  x1 = c(h[1], tt[1], h_hat[1], h[1]),
  y1 = c(h[2], tt[2], h_hat[2], h[2]),
  label = c("h (outcome)", "t (regressor)", "h-hat (fitted)", "e (residual)"),
  color = c("black", "gray50", "forestgreen", "tomato")
)

slope_t <- tt[2] / tt[1]

ggplot() +
  geom_abline(intercept = 0, slope = slope_t, linetype = "dotted", alpha = 0.3) +
  geom_segment(data = df_arrows,
               aes(x = x0, y = y0, xend = x1, yend = y1, color = label),
               arrow = arrow(length = unit(0.15, "inches")),
               linewidth = 1.1) +
  scale_color_manual(values = c("h (outcome)" = "black",
                                "t (regressor)" = "gray50",
                                "h-hat (fitted)" = "forestgreen",
                                "e (residual)" = "tomato"),
                     name = "") +
  coord_fixed(xlim = c(-1, 5), ylim = c(-1, 6)) +
  labs(x = "Observation 1", y = "Observation 2",
       title = "OLS finds the closest point on the line of t to h") +
  theme_minimal()
```

The green vector ($\hat{\mathbf{h}}$) is the best prediction in the "column space" of $\mathbf{t}$, and the red vector ($\mathbf{e}$) is the part of $\mathbf{h}$ that $\mathbf{t}$ cannot explain. With $n = 100$ observations, these vectors live in $\mathbb{R}^{100}$ and we can't draw them — but the geometry is identical. With multiple regressors, the "line" becomes a plane (or hyperplane), and the projection lands on the closest point in that plane.

## Building the design matrix

The model $y = X\beta + e$ stacks $n$ observations into a matrix. Each row of $X$ is one observation; each column is one variable. The first column is typically all ones (the intercept).

We'll use the Canadian Prestige dataset: the Pineo-Porter prestige score of occupations, predicted by average education (years) and average income (dollars) of workers in each occupation.

```{r}
#| label: design-matrix
data(Prestige)
n <- nrow(Prestige)
K <- 3  # intercept + education + income

X <- cbind(1, Prestige$education, Prestige$income)
y <- Prestige$prestige

dim(X)   # n x K
head(X)
```

The two fundamental products in OLS are $X'X$ (a $K \times K$ matrix) and $X'y$ (a $K \times 1$ vector):

```{r}
#| label: xtx-xty
XtX <- crossprod(X)       # K x K: t(X) %*% X
XtX

Xty <- crossprod(X, y)    # K x 1: t(X) %*% y
Xty
```

$X'X$ encodes the relationships among the regressors. The diagonal holds $\sum X_k^2$ for each variable; the off-diagonals hold $\sum X_j X_k$. Dividing by $n$ gives the sample second-moment matrix.

## Bivariate OLS: the formula connection

Before the matrix derivation, recall the bivariate OLS formula: $\hat\beta_1 = \text{Cov}(X, Y)/\text{Var}(X)$. This is the sample analogue of the BLP coefficient from Chapter 2. Let's verify it matches the matrix formula using just education as a predictor:

```{r}
#| label: bivariate-by-hand
educ <- Prestige$education

# Formula approach
beta1_formula <- cov(educ, y) / var(educ)
beta0_formula <- mean(y) - beta1_formula * mean(educ)

# Matrix approach (2x2 system)
X_biv <- cbind(1, educ)
beta_biv <- solve(crossprod(X_biv), crossprod(X_biv, y))

# lm() approach
beta_lm <- coef(lm(prestige ~ education, data = Prestige))

cbind(formula = c(beta0_formula, beta1_formula),
      matrix = beta_biv,
      lm = beta_lm)
```

All three give the same answer. The matrix formula $\hat\beta = (X'X)^{-1}X'y$ generalizes the bivariate $\text{Cov}/\text{Var}$ formula to any number of regressors.

## Deriving OLS with matrix calculus {#sec-ols-derivation}

The sum of squared errors in matrix form is:

$$\text{SSE}(\beta) = (y - X\beta)'(y - X\beta) = \underbrace{y'y}_{\text{constant}} - \underbrace{2y'X\beta}_{\text{linear}} + \underbrace{\beta'X'X\beta}_{\text{quadratic}}$$ {#eq-ols}

Let's build each piece in R and verify the expansion:

```{r}
#| label: sse-expansion
beta_test <- c(0, 1, 0.001)  # an arbitrary beta to test

# Direct computation
sse_direct <- as.numeric(crossprod(y - X %*% beta_test))

# Expanded form
piece1 <- as.numeric(crossprod(y))                         # y'y
piece2 <- as.numeric(2 * crossprod(y, X %*% beta_test))    # 2y'Xbeta
piece3 <- as.numeric(t(beta_test) %*% XtX %*% beta_test)  # beta'X'Xbeta

c(direct = sse_direct, expanded = piece1 - piece2 + piece3)
```

Setting $\partial \text{SSE}/\partial \beta = -2X'y + 2X'X\hat\beta = 0$ gives the **normal equations**:

$$X'X\hat\beta = X'y$$ {#eq-normal-equations}

Solving with `solve()`:

```{r}
#| label: ols-solve
# solve(A, b) solves the system Ax = b — better than solve(A) %*% b
beta_hat <- solve(XtX, Xty)
beta_hat

# lm() gives the same thing
coef(lm(prestige ~ education + income, data = Prestige))
```

Note: `solve(A, b)` is preferred over `solve(A) %*% b` — it avoids computing the full inverse, which is slower and less numerically stable.

::: {#thm-ols}
## The OLS Estimator
The OLS estimator $\hat\beta = (X'X)^{-1}X'y$ is the unique minimizer of $\text{SSE}(\beta) = (y - X\beta)'(y - X\beta)$ when $X'X$ is positive definite.
:::

### What `lm()` actually computes {#sec-qr}

We derived $\hat\beta$ by solving the normal equations $X'X\hat\beta = X'y$, and the code above uses `solve()` to do exactly that. But `lm()` takes a different route: it never forms $X'X$. Instead it uses a **QR decomposition**.

**Why avoid the normal equations?** Every matrix has a condition number $\kappa$ that measures how close it is to singular. Forming $X'X$ *squares* the condition number: if $\kappa(X) = 10^4$, then $\kappa(X'X) = 10^8$, and with ~16 digits of floating-point precision you've lost half your significant digits. For well-conditioned data this doesn't matter, but with near-collinear regressors it can produce garbage.

**The QR approach.** Factor $X = QR$ where $Q$ is $n \times K$ with orthonormal columns ($Q'Q = I_K$) and $R$ is $K \times K$ upper triangular. Then:

$$X'X\hat\beta = X'y \;\;\Longrightarrow\;\; R'Q'QR\hat\beta = R'Q'y \;\;\Longrightarrow\;\; R\hat\beta = Q'y$$

The system $R\hat\beta = Q'y$ is upper triangular, so it's solved by **back-substitution** — starting from the last equation (one unknown) and working up. No matrix inversion needed.

R uses **pivoted Householder reflections** to compute the QR factorization (via LAPACK's DGEQP3). Householder reflections are orthogonal transformations that zero out entries below the diagonal one column at a time — they are more numerically stable than the classical Gram-Schmidt process. The "pivoted" part means R reorders columns to put the most linearly independent ones first, which helps detect rank deficiency (this is how `lm()` drops collinear variables automatically).

```{r}
#| label: qr-decomposition
# lm() stores the QR decomposition
mod <- lm(prestige ~ education + income, data = Prestige)

# Extract Q and R
qr_obj <- qr(X)
Q <- qr.Q(qr_obj)
R <- qr.R(qr_obj)

# Q has orthonormal columns
round(crossprod(Q), 10)

# R is upper triangular
R
```

```{r}
#| label: qr-solve
# Solve R * beta = Q'y by back-substitution
beta_qr <- backsolve(R, crossprod(Q, y))

# Compare all three approaches
cbind(normal_eq = beta_hat, QR = beta_qr, lm = coef(mod))
```

All three agree. The condition number tells us why QR is safer:

```{r}
#| label: condition-numbers
kappa_X <- kappa(X)
kappa_XtX <- kappa(XtX)

c(kappa_X = kappa_X, kappa_XtX = kappa_XtX, ratio = kappa_XtX / kappa_X)
```

The ratio is approximately $\kappa(X)$ — forming $X'X$ squares the condition number, as expected. For this well-conditioned dataset the normal equations work fine, but `lm()` uses QR as a safeguard.

### The second-order condition

The second derivative of $\text{SSE}$ is $2X'X$. This is a minimum when $X'X$ is positive definite — all eigenvalues are positive:

```{r}
#| label: second-order
eigen(XtX)$values
```

All positive, confirming positive definiteness. If any eigenvalue were zero, $X'X$ would be singular and `solve()` would fail.

## What collinearity does to $X'X$

When a column of $X$ is a linear combination of others, $X'X$ loses rank:

```{r}
#| label: collinearity
# Add a redundant column: income2 = 2 * income
X_bad <- cbind(X, 2 * Prestige$income)

XtX_bad <- crossprod(X_bad)
det(XtX_bad)          # essentially zero
eigen(XtX_bad)$values  # last eigenvalue collapses
```

In practice, *near*-collinearity (very small but nonzero eigenvalues) inflates standard errors without crashing `solve()`. The **condition number** — ratio of largest to smallest eigenvalue — measures how close to singular:

```{r}
#| label: condition-number
evals <- eigen(XtX)$values
c(largest = evals[1], smallest = evals[K], condition = evals[1] / evals[K])
```

R's `lm()` handles exact collinearity by dropping the redundant column:

```{r}
#| label: lm-collinearity
coef(lm(prestige ~ education + income + I(2 * income), data = Prestige))
```

::: {.callout-warning}
## Near-Collinearity Inflates Standard Errors
When columns of $X$ are nearly linearly dependent, $X'X$ has a near-zero eigenvalue, making $(X'X)^{-1}$ very large. This inflates the variance of $\hat\beta$ without causing `solve()` to fail — standard errors balloon silently. Check the condition number of $X'X$ to detect this.
:::

## The projection matrix {#sec-projection-matrix}

The projection matrix $P = X(X'X)^{-1}X'$ maps any $n$-vector onto the column space of $X$. In two dimensions (our earlier example), it projected $\mathbf{h}$ onto the line of $\mathbf{t}$. With $K = 3$ regressors, it projects $\mathbf{y}$ onto a 3-dimensional subspace of $\mathbb{R}^n$.

```{r}
#| label: projection-matrix
P <- X %*% solve(XtX) %*% t(X)
dim(P)  # n x n

mod <- lm(prestige ~ education + income, data = Prestige)
```

::: {#def-projection-matrix}
## Projection (Hat) Matrix
The projection matrix $P = X(X'X)^{-1}X'$ maps any $n$-vector onto the column space of $X$. It is symmetric ($P' = P$) and idempotent ($P^2 = P$), with eigenvalues in $\{0, 1\}$ and $\text{tr}(P) = K$.
:::

Every property of $P$ corresponds to a matrix operation:

```{r}
#| label: projection-properties
# P*y = fitted values
all.equal(as.vector(P %*% y), as.numeric(fitted(mod)))

# Symmetric: t(P) = P
all.equal(t(P), P)

# Idempotent: P %*% P = P (projecting twice = projecting once)
all.equal(P %*% P, P)

# P*X = X (X is already in its own column space)
all.equal(P %*% X, X, check.attributes = FALSE)
```

### What idempotency means for eigenvalues

If $Pv = \lambda v$, then $P^2 v = \lambda^2 v$. But $P^2 = P$, so $\lambda^2 = \lambda$, which means $\lambda \in \{0, 1\}$:

```{r}
#| label: projection-eigenvalues
eig_P <- eigen(P)$values
table(round(eig_P, 10))
```

$K$ eigenvalues equal 1 (the column space of $X$) and $n - K$ equal 0 (the null space). The trace counts the 1s:

```{r}
#| label: projection-trace
c(trace_P = tr(P), K = K)
```

### Projection onto the intercept

The simplest projection is onto a column of ones: $P_1 = \mathbf{1}(\mathbf{1}'\mathbf{1})^{-1}\mathbf{1}' = \frac{1}{n}\mathbf{1}\mathbf{1}'$. This projects every observation onto the sample mean:

```{r}
#| label: intercept-projection
ones <- rep(1, n)
P1 <- outer(ones, ones) / n  # outer product: 1*1' / n

# P1 * y = sample mean for every observation
all.equal(as.vector(P1 %*% y), rep(mean(y), n))
```

Every additional regressor refines this baseline: the full $P$ starts from the mean and adds the directions explained by the other columns of $X$.

## The annihilator matrix

The annihilator $M = I_n - P$ projects onto the orthogonal complement — the part of $y$ that $X$ cannot explain:

```{r}
#| label: annihilator-matrix
M <- diag(n) - P

# M*y = residuals
all.equal(as.vector(M %*% y), as.numeric(resid(mod)))

# Idempotent and symmetric
all.equal(M %*% M, M)
all.equal(t(M), M)

# M annihilates X: M*X = 0
max(abs(M %*% X))
```

::: {#def-annihilator}
## Annihilator Matrix
The annihilator $M = I_n - P$ projects onto the orthogonal complement of the column space of $X$. It satisfies $MX = 0$ (annihilates $X$), is idempotent and symmetric, and has $\text{tr}(M) = n - K$.
:::

Eigenvalues are complementary to $P$: $n - K$ ones and $K$ zeros:

```{r}
#| label: annihilator-eigenvalues
c(trace_M = tr(M), n_minus_K = n - K)
```

The **demeaning matrix** $M_1 = I - P_1$ is a special case — it subtracts the mean:

```{r}
#| label: demeaning
M1 <- diag(n) - P1
all.equal(as.vector(M1 %*% y), as.numeric(y - mean(y)))
```

## Residuals vs. disturbances

The true model is $y = X\beta + e$ where $e$ is unobservable. The residuals $\hat{e} = My$ relate to the disturbances through:

$$\hat{e} = My = M(X\beta + e) = \underbrace{MX}_{= 0}\beta + Me = Me$$

Let's simulate to see this. We know $\beta$ and $e$ because we generate the data:

```{r}
#| label: resid-vs-disturbance
set.seed(307)
n_sim <- 100
K_sim <- 2
X_sim <- cbind(1, rnorm(n_sim))
beta_true <- c(2, 3)
e_true <- rnorm(n_sim, sd = 2)
y_sim <- X_sim %*% beta_true + e_true

# Build M for this design
P_sim <- X_sim %*% solve(crossprod(X_sim)) %*% t(X_sim)
M_sim <- diag(n_sim) - P_sim

# Residuals = M * disturbances
e_hat <- as.vector(M_sim %*% y_sim)
all.equal(e_hat, as.vector(M_sim %*% e_true))

# Residuals have smaller variance — M zeroes out K dimensions
c(var_disturbances = var(e_true), var_residuals = var(e_hat))
```

## Estimating $\sigma^2$: the trace trick

The natural estimator $\hat\sigma^2 = \hat{e}'\hat{e}/n$ is biased downward because $\hat{e}'\hat{e} = e'Me \leq e'e$ ($M$ is positive semi-definite). The unbiased estimator divides by $n - K$.

The proof is a chain of matrix operations. Every step translates to R:

```{r}
#| label: trace-trick
# Step 1: e'Me is a scalar = its own trace
scalar_form <- as.numeric(t(e_true) %*% M_sim %*% e_true)
trace_form <- tr(M_sim %*% tcrossprod(e_true))  # tr(M * ee')

c(scalar = scalar_form, trace = trace_form)

# Step 2: E[ee'] = sigma^2 * I, so E[tr(Mee')] = sigma^2 * tr(M)
# tr(M) = n - K, so E[e'hat * e'hat] = sigma^2 * (n - K)
c(trace_M = tr(M_sim), n_minus_K = n_sim - K_sim)
```

::: {.callout-note}
## The $n - K$ Divisor
The unbiased variance estimator divides by $n - K$ (not $n$) because the residuals live in an $(n - K)$-dimensional subspace. The $K$ "lost" dimensions are consumed by estimating $\hat\beta$. This is the matrix version of Bessel's correction.
:::

This is why the unbiased estimator is $s_e^2 = \hat{e}'\hat{e}/(n-K)$:

```{r}
#| label: error-variance
# Back to the Prestige data
e_hat_prestige <- resid(mod)

sigma2_biased <- as.numeric(crossprod(e_hat_prestige)) / n
sigma2_unbiased <- as.numeric(crossprod(e_hat_prestige)) / (n - K)

c(biased = sigma2_biased,
  unbiased = sigma2_unbiased,
  R_sigma2 = sigma(mod)^2)
```

The underestimation shows up in the eigenvalues of $M$: $n - K$ eigenvalues are 1, and $K$ are 0. The residuals live in an $(n-K)$-dimensional subspace:

```{r}
#| label: m-eigenvalues
eig_M <- round(eigen(M)$values, 10)
c(eigenvalues_equal_1 = sum(eig_M == 1),
  eigenvalues_equal_0 = sum(eig_M == 0))
```

## Variance of $\hat\beta$: building $s_e^2(X'X)^{-1}$

Under homoskedasticity, $\text{Var}(\hat\beta|X) = \sigma^2(X'X)^{-1}$. Each piece is a matrix operation:

```{r}
#| label: vcov-construction
# Step 1: (X'X)^{-1}
XtX_inv <- solve(XtX)
XtX_inv

# Step 2: multiply by s_e^2
vcov_manual <- sigma(mod)^2 * XtX_inv

# Step 3: compare to R
all.equal(vcov_manual, vcov(mod), check.attributes = FALSE)
```

Standard errors are the square roots of the diagonal:

```{r}
#| label: standard-errors
se_manual <- sqrt(diag(vcov_manual))
se_R <- coef(summary(mod))[, "Std. Error"]

cbind(manual = se_manual, R = se_R)
```

### Why $(X'X)^{-1}$ determines precision

The eigenvalues of $(X'X)^{-1}$ are the reciprocals of those of $X'X$. Large eigenvalues of $X'X$ (strong signal) become small eigenvalues of $(X'X)^{-1}$ (precise estimates). Near-collinearity creates a tiny eigenvalue in $X'X$, which inflates variance:

```{r}
#| label: xtx-inv-eigenvalues
eig_XtX <- eigen(XtX)$values
eig_inv <- eigen(XtX_inv)$values

cbind(XtX = eig_XtX, XtX_inv = eig_inv, product = eig_XtX * eig_inv)
```

The products are all 1: the eigenvalues invert exactly.

## Application: the Prestige regression

Let's interpret the full regression. Education and income both predict occupational prestige:

```{r}
#| label: prestige-application
summary(mod)
```

The coefficient on education (`r round(coef(mod)[2], 1)`) says: holding income constant, one additional year of average education is associated with about `r round(coef(mod)[2], 1)` points more prestige. The coefficient on income (`r round(coef(mod)[3], 4)`) is small in magnitude because income is in dollars — a \$1,000 increase predicts about `r round(coef(mod)[3] * 1000, 1)` points.

Let's see which occupations the model fits well and poorly, using the projection and annihilator:

```{r}
#| label: prestige-residuals
Prestige$fitted <- as.vector(P %*% y)
Prestige$resid <- as.vector(M %*% y)

# Largest positive residuals: more prestige than education+income predict
head(Prestige[order(-Prestige$resid), c("education", "income", "prestige", "fitted", "resid")], 5)

# Largest negative residuals: less prestige than predicted
head(Prestige[order(Prestige$resid), c("education", "income", "prestige", "fitted", "resid")], 5)
```

```{r}
#| label: prestige-residual-plot
ggplot(Prestige, aes(fitted, resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text(data = Prestige[abs(Prestige$resid) > 15, ],
            aes(label = rownames(Prestige)[abs(Prestige$resid) > 15]),
            hjust = -0.1, size = 3) +
  labs(x = "Fitted values (Py)", y = "Residuals (My)",
       title = "Prestige: fitted vs. residuals") +
  theme_minimal()
```

## ANOVA as inner products

The decomposition $y = \hat{y} + \hat{e}$ is orthogonal. In matrix terms, $\hat{y}'\hat{e} = (Py)'(My) = y'PMy = 0$:

```{r}
#| label: orthogonality
as.numeric(crossprod(fitted(mod), resid(mod)))
```

After centering, the inner products give sums of squares:

```{r}
#| label: anova-decomposition
# SST = ||M1 * y||^2  (total variation around the mean)
SST <- as.numeric(crossprod(M1 %*% y))

# SSR = ||(P - P1) * y||^2  (variation explained by regressors beyond the mean)
SSR <- as.numeric(crossprod(fitted(mod) - mean(y)))

# SSE = ||M * y||^2  (unexplained variation)
SSE <- as.numeric(crossprod(resid(mod)))

c(SST = SST, SSR_plus_SSE = SSR + SSE)

# Cross term is zero when X includes a constant
as.numeric(crossprod(fitted(mod) - mean(y), resid(mod)))
```

## $R^2$

$R^2 = \text{SSR}/\text{SST} = 1 - \text{SSE}/\text{SST}$:

```{r}
#| label: r-squared
c(R2 = SSR / SST,
  R2_alt = 1 - SSE / SST,
  R_reports = summary(mod)$r.squared)
```

Adding regressors can only increase $R^2$, even if the variable is noise. The adjusted $R^2$ penalizes for $K$:

```{r}
#| label: adjusted-r2
set.seed(42)
Prestige$noise <- rnorm(n)
mod_noise <- lm(prestige ~ education + income + noise, data = Prestige)

c(R2_original = summary(mod)$r.squared,
  R2_with_noise = summary(mod_noise)$r.squared,
  adj_R2_original = summary(mod)$adj.r.squared,
  adj_R2_with_noise = summary(mod_noise)$adj.r.squared)
```

Raw $R^2$ ticks up; adjusted $R^2$ drops — correctly penalizing the useless variable.

$R^2$ measures descriptive fit, not causal validity. Typical values: cross-sectional micro data $\approx 0.2$--$0.4$, aggregate time series $\approx 0.7$--$0.9$.

## Naming conventions: a warning

Different textbooks use SSE and SSR with opposite meanings. In this course (following Hansen), SSR is "regression" (explained) and SSE is "error" (unexplained). Some texts reverse these. The math is always $\text{SST} = \text{explained} + \text{unexplained}$.

## Summary

The OLS estimator is a sequence of matrix operations:

| Math | R | What it does |
|------|---|-------------|
| $X'X$ | `crossprod(X)` | Gram matrix of regressors |
| $(X'X)^{-1}$ | `solve(crossprod(X))` | Inverse |
| $\hat\beta = (X'X)^{-1}X'y$ | `solve(crossprod(X), crossprod(X, y))` | OLS coefficients (normal equations) |
| $X = QR$ | `qr(X)` | QR decomposition (what `lm()` uses) |
| $R\hat\beta = Q'y$ | `backsolve(qr.R(qr(X)), crossprod(qr.Q(qr(X)), y))` | OLS via back-substitution |
| $P = X(X'X)^{-1}X'$ | `X %*% solve(crossprod(X)) %*% t(X)` | Projection (hat matrix) |
| $M = I - P$ | `diag(n) - P` | Annihilator |
| $\text{tr}(M)$ | `sum(diag(M))` | Degrees of freedom ($n - K$) |
| eigenvalues of $P$ | `eigen(P)$values` | All 0 or 1 |
| $s_e^2(X'X)^{-1}$ | `sigma(mod)^2 * solve(crossprod(X))` | Variance-covariance of $\hat\beta$ |
| $\text{SE}(\hat\beta_k)$ | `sqrt(diag(vcov(mod)))` | Standard errors |

Key facts:

- OLS is **projection**: $\hat{y} = Py$ is the closest point to $y$ in the column space of $X$, and $\hat{e} = My$ is the orthogonal residual.
- $P$ and $M$ are **symmetric** and **idempotent**, with eigenvalues in $\{0, 1\}$.
- $\text{tr}(P) = K$ and $\text{tr}(M) = n - K$ count dimensions.
- The **trace trick** proves $s_e^2$ is unbiased: $\mathbb{E}[e'Me] = \sigma^2 \text{tr}(M) = \sigma^2(n-K)$.
- Eigenvalues of $(X'X)^{-1}$ are reciprocals of those of $X'X$: near-collinearity inflates variance.
- `lm()` uses **QR decomposition** (pivoted Householder reflections), not the normal equations — this avoids squaring the condition number of $X$.
- **Regression to the mean** is a consequence of $\hat\beta_1 < 1$ when $|\rho| < 1$.

Next: [Sensitivity and Leverage](ch04-sensitivity.qmd) — the Frisch-Waugh-Lovell theorem, partial $R^2$, and influential observations.
