---
title: "9. Hypothesis Testing"
subtitle: "The F-test, the test trinity, multiple testing, and power"
---

Chapter 8 gave us sandwich standard errors and bootstrap confidence intervals. This chapter puts them to work: we test hypotheses, build confidence regions, guard against false discoveries, and ask whether our sample is large enough to detect effects we care about.

The emphasis is on *doing* --- running tests in R, interpreting the output correctly, and avoiding common pitfalls. We organize around applied questions:

1. **How do I test whether a coefficient is zero?** (t-tests, robust Wald)
2. **How do I test multiple restrictions at once?** (F-test, joint Wald)
3. **Why Wald, not LR or classical F, is the default** (the trinity under misspecification)
4. **How do I get confidence sets for nonlinear parameters?** (test inversion, Fieller)
5. **What if I test many hypotheses?** (Bonferroni, Holm)
6. **How large a sample do I need?** (power analysis)

We use the `Salaries` dataset from `carData` throughout: salaries of 397 U.S. professors, with rank, discipline, years since PhD, years of service, and sex.

```{r}
#| label: setup
#| message: false
library(ggplot2)
library(sandwich)
library(lmtest)
library(estimatr)
library(car)
library(carData)
options(digits = 4)

data(Salaries)
```

## Testing a single coefficient

We start with a wage equation and ask: controlling for rank, discipline, and experience, does sex predict salary?

```{r}
#| label: base-model
mod <- lm(salary ~ rank + discipline + yrs.since.phd + yrs.service + sex,
          data = Salaries)
summary(mod)
```

The `summary()` output gives t-statistics and p-values, but these assume homoskedasticity. The robust version uses sandwich standard errors:

```{r}
#| label: robust-t
coeftest(mod, vcov. = vcovHC(mod, type = "HC1"))
```

These are the same tests, built from the same Wald statistic $W = (\hat{\beta}_j / \widehat{\text{se}}_j)^2$. The only difference is the denominator. When the two disagree, trust the robust version --- it doesn't assume $\text{Var}(e_i \mid X_i)$ is constant.

You can also use `linearHypothesis()` for the same test, which makes the Wald structure explicit:

```{r}
#| label: single-linearHypothesis
linearHypothesis(mod, "sexMale = 0", vcov. = vcovHC(mod, type = "HC1"))
```

This tests $H_0: \beta_{\text{Male}} = 0$ against $H_1: \beta_{\text{Male}} \neq 0$ using the robust covariance matrix. The Chisq column is the Wald statistic; divide by 1 (one restriction) and you get the squared t-statistic.

### Testing against a non-zero value

Suppose theory predicts a gender gap of \$5,000. The null is $H_0: \beta_{\text{Male}} = 5000$:

```{r}
#| label: test-nonzero
linearHypothesis(mod, "sexMale = 5000", vcov. = vcovHC(mod, type = "HC1"))
```

The test is not significant if the confidence interval from Chapter 8 already contains \$5,000. Test and CI carry exactly the same information --- they are duals.


## Testing multiple restrictions: the F-test

Individual t-tests tell you about one coefficient at a time. But sometimes you need a *joint* test. Does rank matter at all? That means testing two coefficients simultaneously (since rank is a three-level factor with two dummies):

$$H_0: \beta_{\text{AssocProf}} = \beta_{\text{Prof}} = 0$$

### By hand: comparing two regressions

The F-test compares the fit of an **unrestricted** model to a **restricted** model that imposes $H_0$:

$$F = \frac{(\text{SSE}_R - \text{SSE}_U) / q}{\text{SSE}_U / (n - k)}$$

where $q$ is the number of restrictions.

```{r}
#| label: f-test-manual
## Unrestricted model (full)
mod_U <- mod

## Restricted model: drop rank
mod_R <- lm(salary ~ discipline + yrs.since.phd + yrs.service + sex,
            data = Salaries)

## Manual F-stat
SSE_U <- sum(resid(mod_U)^2)
SSE_R <- sum(resid(mod_R)^2)
q <- 2                              # two restrictions
n <- nobs(mod_U)
k <- length(coef(mod_U))

F_stat <- ((SSE_R - SSE_U) / q) / (SSE_U / (n - k))
p_val  <- 1 - pf(F_stat, q, n - k)

cat("F =", round(F_stat, 2), "  df1 =", q, "  df2 =", n - k,
    "  p =", format.pval(p_val), "\n")
```

### The easy way: `anova()` and `linearHypothesis()`

```{r}
#| label: f-test-anova
## Nested model comparison
anova(mod_R, mod_U)
```

```{r}
#| label: f-test-linearHypothesis
## Same test via linearHypothesis (with robust SEs)
linearHypothesis(mod, c("rankAssocProf = 0", "rankProf = 0"),
                 vcov. = vcovHC(mod, type = "HC1"))
```

Note that `anova()` uses the classical (homoskedastic) F-test, while `linearHypothesis()` with a robust `vcov.` gives the heteroskedasticity-robust Wald test. In this example they agree; with strongly heteroskedastic data they might not.

### Equality constraints

Does the return to years since PhD equal the return to years of service? This is $H_0: \beta_{\text{yrs.phd}} = \beta_{\text{yrs.service}}$:

```{r}
#| label: equality-test
linearHypothesis(mod, "yrs.since.phd = yrs.service",
                 vcov. = vcovHC(mod, type = "HC1"))
```

This works because `linearHypothesis()` can test *any* linear restriction $\mathbf{R}\boldsymbol{\beta} = \mathbf{r}$.  Under the hood it constructs the restriction matrix $\mathbf{R}$ and computes:
$$W = (\mathbf{R}\hat{\boldsymbol{\beta}} - \mathbf{r})' [\mathbf{R}\hat{\mathbf{V}}\mathbf{R}']^{-1} (\mathbf{R}\hat{\boldsymbol{\beta}} - \mathbf{r}) \;\xrightarrow{d}\; \chi^2_q$$ {#eq-wald}

::: {#thm-wald}
## Wald Statistic
For testing $H_0: R\beta = r$, the Wald statistic is $W = (R\hat\beta - r)'[R\hat{V}R']^{-1}(R\hat\beta - r) \xrightarrow{d} \chi^2_q$ under $H_0$, where $q$ is the number of restrictions. With robust $\hat{V}$, the test is valid under heteroskedasticity.
:::


## The test trinity under misspecification

Three classical tests attack the same null $H_0: R\beta = r$:

| Test | Based on | Needs to believe |
|------|----------|------------------|
| **Wald** | distance from $\hat\beta$ to the null | only the mean function (plus consistent $\hat V$) |
| **Score / LM** | slope of the objective at the restricted estimator | only the mean function (plus consistent "meat" matrix) |
| **LR** | height difference $\log L_U - \log L_R$ | the full likelihood |

Two things about this table are worth dwelling on.

**LR requires the likelihood.** Under homoskedastic normal errors the classical F-test equals the LR statistic up to a monotone transformation, and the whole trinity is pedagogically unified. Once errors are heteroskedastic or the likelihood is otherwise misspecified, $-2(\log L_R - \log L_U)$ no longer converges to $\chi^2_q$ --- it converges to a weighted sum of $\chi^2_1$ random variables with unknown weights. There are Satterthwaite-style corrections ("quasi-LR") but in practice applied econometrics has simply dropped LR as a default and uses Wald with sandwich variance. The same logic rules out the classical F from `anova()`, which is the homoskedastic LR for linear models.

**Wald and Score have clean robust versions.** Swap the classical variance for the sandwich:
$$W = (R\hat\beta - r)'\,[R\,\hat V_{\text{HC}}\, R']^{-1}\,(R\hat\beta - r) \xrightarrow{d} \chi^2_q$$
$$LM = \Big(\textstyle\sum s_i\Big)'\Big(\textstyle\sum s_i s_i'\Big)^{-1}\Big(\textstyle\sum s_i\Big) \xrightarrow{d} \chi^2_q$$
The Wald is `linearHypothesis(mod, ..., vcov = vcovHC(mod, type = "HC2"))`. The robust LM comes up most often not as a head-to-head rival to Wald but as the engine of **specification tests** --- Breusch–Pagan for heteroskedasticity, White's information-matrix test, and (Chapter 11) the GMM overidentification test are all score tests.

### When do you still use the classical F?

One genuinely remaining niche: when you *do* believe errors are normal and homoskedastic, the classical F has an **exact** $F_{q,\,n-K}$ distribution in finite samples, while the robust Wald is only asymptotic. At $n = 20$ with trustworthy homoskedasticity, the classical F's exact critical values can be better-calibrated than $\chi^2_q$. By $n = 200$ this advantage vanishes.

### A concrete comparison

```{r}
#| label: wald-vs-anova
## Joint test: rank has no effect (q = 2 restrictions)
## Classical F (homoskedastic)
anova(mod_R, mod_U)$F[2]

## Robust Wald (heteroskedasticity-robust)
linearHypothesis(mod_U, c("rankAssocProf = 0", "rankProf = 0"),
                 vcov = vcovHC(mod_U, type = "HC2"),
                 test = "F")$F[2]
```

The two statistics answer slightly different questions: classical F asks "under the homoskedastic-normal model, does fit deteriorate?"; robust Wald asks "given the estimated sandwich covariance, is $R\hat\beta$ far from $r$?" When errors are actually homoskedastic they give nearly identical numbers. Otherwise, trust the robust version.

### Where Score tests live

Score tests show up throughout the course, mostly not labeled as such:

- **Breusch–Pagan** (ch05): score test against $H_0: \text{Var}(e_i \mid X_i)$ = constant.
- **RESET** (ch04): score test against $H_0:$ no omitted nonlinear terms.
- **Hausman** (ch12): comparison of two estimators is equivalent to a score-type test.
- **GMM overidentification / $J$-test** (ch11): the canonical modern score test --- are extra moment conditions satisfied at the efficient estimator?

The common thread: score tests only require the **null** fit. You estimate under $H_0$ and ask whether the remaining moment conditions or specification implications are satisfied at that estimate. Two reasons this is the right tool:

- **There may be no alternative model to fit.** In GMM overidentification you have more moment conditions than parameters. There is no larger model that uses them all as free structure --- the extras are just claims about the true $\theta$. The score test asks: at the estimate pinned down by $K$ moments, are the remaining $L - K$ moments close to zero?
- **The alternative is too broad to parametrize.** Breusch–Pagan tests against "some form of heteroskedasticity" without naming which form. You don't want to fit every possible heteroskedastic model; you want to check whether the residuals from the homoskedastic fit betray any systematic pattern.

In both cases, Wald is not available --- there is nothing unrestricted to compare against. The score test is what you reach for.


## Confidence regions and test inversion

A 95% confidence interval is the set of values $\theta_0$ that would *not be rejected* at the 5% level:

$$\hat{C} = \{\theta_0 : |T(\theta_0)| \leq 1.96\}$$

This is **test inversion**. For a single linear coefficient, test inversion gives the familiar $\hat{\beta} \pm 1.96 \cdot \widehat{\text{se}}$. But for *nonlinear* functions of parameters, test inversion is more reliable than the delta method.

::: {.callout-note}
## Test Inversion Gives Confidence Sets
A 95% confidence set is exactly the set of parameter values not rejected at the 5% level. For linear parameters, this gives the familiar $\hat\beta \pm 1.96 \cdot \text{SE}$. For nonlinear functions, Fieller's method (test inversion over a grid) is more reliable than the delta method.
:::

### Confidence ellipses for two coefficients

For two parameters tested jointly, the confidence region is an **ellipse** --- the set of $(\beta_1, \beta_2)$ values consistent with the Wald test:

```{r}
#| label: confidence-ellipse
#| fig-width: 7
#| fig-height: 5.5
## Joint confidence region for the two experience variables
confidenceEllipse(mod, which.coef = c("yrs.since.phd", "yrs.service"),
                  vcov. = vcovHC(mod, type = "HC1"),
                  main = "95% joint confidence region",
                  xlab = "Coefficient on yrs.since.phd",
                  ylab = "Coefficient on yrs.service",
                  col = "steelblue", lwd = 2)
abline(h = 0, lty = 2, col = "gray50")
abline(v = 0, lty = 2, col = "gray50")
## Add the 45-degree line for equality
abline(a = 0, b = 1, lty = 3, col = "coral", lwd = 1.5)
legend("topright", legend = c("95% joint region", "equality line"),
       col = c("steelblue", "coral"), lty = c(1, 3), lwd = c(2, 1.5))
```

The dashed lines at zero show the individual null values. The coral line shows where the two coefficients are equal. If the ellipse crosses a line, the corresponding null cannot be rejected.

### Test inversion for a nonlinear parameter: Fieller's method

Suppose we want a confidence interval for the **ratio** $\theta = \beta_{\text{Prof}} / \beta_{\text{AssocProf}}$ --- how many times larger is the full professor salary premium than the associate professor premium? The delta method (Chapter 8) gives a quick CI, but it can be unreliable when the denominator is imprecisely estimated.

**Fieller's trick**: rewrite $\theta = \beta_1 / \beta_2$ as the linear restriction $\beta_1 - \theta \beta_2 = 0$, then *invert* the Wald test over a grid of $\theta$ values:

```{r}
#| label: fieller-inversion
b <- coef(mod)
V <- vcovHC(mod, type = "HC1")

## Delta method CI for comparison
dm <- deltaMethod(mod, "rankProf / rankAssocProf",
                  vcov. = vcovHC(mod, type = "HC1"))
dm_ci <- c(dm$Estimate - 1.96 * dm$SE, dm$Estimate + 1.96 * dm$SE)

## Grid search: invert the Wald test
theta_grid <- seq(0, 8, by = 0.01)
idx_prof <- which(names(b) == "rankProf")
idx_assoc <- which(names(b) == "rankAssocProf")

in_CI <- sapply(theta_grid, function(th) {
  r_vec <- numeric(length(b))
  r_vec[idx_prof] <- 1
  r_vec[idx_assoc] <- -th
  num <- (sum(r_vec * b))^2
  den <- t(r_vec) %*% V %*% r_vec
  num / den <= qchisq(0.95, 1)
})
fieller_ci <- range(theta_grid[in_CI])

cat("Point estimate:  ", round(dm$Estimate, 3), "\n")
cat("Delta method CI: ", round(dm_ci, 3), "\n")
cat("Fieller CI:      ", round(fieller_ci, 3), "\n")
```

Here the two CIs are similar because the denominator ($\beta_{\text{AssocProf}}$) is precisely estimated. The ratio tells us that the full professor premium is about 3.5 times the associate professor premium.

```{r}
#| label: fieller-plot
#| fig-width: 8
#| fig-height: 3.5
## Plot the Wald statistic over theta
wald_vals <- sapply(theta_grid, function(th) {
  r_vec <- numeric(length(b))
  r_vec[idx_prof] <- 1
  r_vec[idx_assoc] <- -th
  num <- (sum(r_vec * b))^2
  den <- c(t(r_vec) %*% V %*% r_vec)
  num / den
})

df_fieller <- data.frame(theta = theta_grid, W = wald_vals)
ggplot(df_fieller, aes(theta, W)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = qchisq(0.95, 1), linetype = "dashed", color = "coral") +
  annotate("text", x = 6.5, y = qchisq(0.95, 1) + 0.5, label = "95% critical value",
           color = "coral") +
  coord_cartesian(ylim = c(0, 15)) +
  labs(title = "Test inversion: Wald statistic as a function of θ₀",
       subtitle = "The Fieller CI is the set of θ₀ values below the dashed line",
       x = expression(theta[0] == beta[Prof] / beta[AssocProf]),
       y = "Wald statistic")
```

### When Fieller and the delta method disagree

The real value of Fieller's method appears when the **denominator is imprecise**. Consider $\theta = \beta_{\text{yrs.phd}} / \beta_{\text{yrs.service}}$ --- the ratio of the two experience coefficients. Both are individually insignificant:

```{r}
#| label: fieller-unbounded
## Experience coefficients are imprecise
cat("yrs.since.phd: estimate =", round(b["yrs.since.phd"]),
    "  SE =", round(sqrt(V["yrs.since.phd", "yrs.since.phd"])), "\n")
cat("yrs.service:   estimate =", round(b["yrs.service"]),
    "  SE =", round(sqrt(V["yrs.service", "yrs.service"])), "\n")

## Delta method gives a finite CI
dm2 <- deltaMethod(mod, "yrs.since.phd / yrs.service",
                   vcov. = vcovHC(mod, type = "HC1"))
cat("\nDelta method CI:", round(c(dm2$Estimate - 1.96 * dm2$SE,
                                  dm2$Estimate + 1.96 * dm2$SE), 2), "\n")

## Fieller: check a wide grid
theta_wide <- seq(-50, 50, by = 0.1)
idx_phd <- which(names(b) == "yrs.since.phd")
idx_svc <- which(names(b) == "yrs.service")
in_CI2 <- sapply(theta_wide, function(th) {
  r_vec <- numeric(length(b))
  r_vec[idx_phd] <- 1
  r_vec[idx_svc] <- -th
  num <- (sum(r_vec * b))^2
  den <- c(t(r_vec) %*% V %*% r_vec)
  num / den <= qchisq(0.95, 1)
})

## Check if CI covers entire grid (unbounded)
if (all(in_CI2)) {
  cat("Fieller CI: (-∞, +∞)  [unbounded!]\n")
} else if (any(in_CI2)) {
  cat("Fieller CI:", range(theta_wide[in_CI2]), "\n")
}
```

The delta method reports a finite CI of width ~1.3, but the Fieller CI is **unbounded** --- the entire real line. This is the correct answer: since $\beta_{\text{yrs.service}}$ is not significantly different from zero, the ratio could be anything. The delta method's finite interval is overconfident.


## Multiple testing {#sec-multiple-testing}

When you look at many coefficients, some will be "significant" by chance.

### Simulation: false discoveries under the null

To make this concrete: generate 20 pure-noise regressors, test each at 5%, and count how many reject:

```{r}
#| label: multiple-testing-sim
set.seed(303)
n <- 200
k_noise <- 20
Y <- rnorm(n)
X_noise <- matrix(rnorm(n * k_noise), nrow = n)
colnames(X_noise) <- paste0("X", 1:k_noise)
dat_noise <- data.frame(Y, X_noise)

mod_noise <- lm(Y ~ ., data = dat_noise)
pvals <- summary(mod_noise)$coefficients[-1, 4]  # drop intercept

cat("Number of regressors with p < 0.05:", sum(pvals < 0.05), "out of", k_noise, "\n")
cat("Significant variables:", paste(names(pvals[pvals < 0.05]), collapse = ", "), "\n")
```

Under the global null, we *expect* $20 \times 0.05 = 1$ false rejection on average. Repeat this many times to see the problem clearly:

```{r}
#| label: fwer-simulation
#| fig-width: 7
#| fig-height: 4
set.seed(42)
B <- 5000
n <- 200
k_noise <- 20

false_rejections <- replicate(B, {
  Y <- rnorm(n)
  X <- matrix(rnorm(n * k_noise), nrow = n)
  pv <- summary(lm(Y ~ X))$coefficients[-1, 4]
  sum(pv < 0.05)
})

cat("P(at least one false rejection):", mean(false_rejections >= 1), "\n")
cat("Expected:", round(1 - (1 - 0.05)^k_noise, 3), "\n")

ggplot(data.frame(rejections = false_rejections), aes(rejections)) +
  geom_bar(fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "coral") +
  labs(title = "False rejections when all 20 nulls are true",
       subtitle = paste0("P(≥1 false rejection) = ",
                         round(mean(false_rejections >= 1), 3)),
       x = "Number of rejections at α = 0.05",
       y = "Frequency")
```

With 20 tests at 5%, you have about a `r round(100 * (1 - (1 - 0.05)^20))`% chance of at least one false rejection. This is the **familywise error rate (FWER)** problem.

::: {.callout-warning}
## Multiple Testing Inflates the Familywise Error Rate
With $k$ independent tests at level $\alpha$, the probability of at least one false rejection is $1 - (1 - \alpha)^k$. With 20 tests at 5%, this exceeds 60%. Use Holm's correction (uniformly more powerful than Bonferroni) to control the FWER.
:::

### Corrections: Bonferroni and Holm

The **Bonferroni correction** rejects the $j$th hypothesis only if $p_j < \alpha / k$. It controls the FWER for any dependence structure:

```{r}
#| label: bonferroni-holm
## Back to our salary model
pvals_salary <- summary(mod)$coefficients[-1, 4]

corrections <- data.frame(
  raw = round(pvals_salary, 4),
  bonferroni = round(p.adjust(pvals_salary, method = "bonferroni"), 4),
  holm = round(p.adjust(pvals_salary, method = "holm"), 4)
)
corrections
```

**Holm's method** is uniformly more powerful than Bonferroni while still controlling the FWER. It works by ordering the p-values from smallest to largest and comparing the $j$th smallest to $\alpha / (k - j + 1)$. Use Holm whenever you would use Bonferroni.

### When to worry about multiple testing

Multiple testing corrections matter when you:

- Examine many coefficients and report the "most significant"
- Try many specifications and present the one that "works"
- Test across subgroups (by age, gender, region, ...)
- Use stepwise selection procedures

They matter *less* when you have a small number of pre-specified hypotheses motivated by theory.


## Power: can you detect the effect?

A test's **power** is the probability it rejects when $H_0$ is false:

$$\text{Power}(\theta) = P[\text{Reject } H_0 \mid \theta \neq \theta_0]$$

Power depends on four things: the true effect size, the standard error, the significance level, and the sample size. In OLS, for a two-sided test of $H_0: \beta = 0$:

$$\delta = \frac{|\beta|}{\text{se}(\hat{\beta})} \approx \frac{|\beta| \cdot \text{sd}(X) \cdot \sqrt{n}}{\sigma_e}$$

Power $\approx \Phi(\delta - z_{\alpha/2}) + \Phi(-\delta - z_{\alpha/2})$, where $\delta$ is the signal-to-noise ratio.

::: {#thm-power}
## Power of a Two-Sided Test
For testing $H_0: \beta = 0$ at level $\alpha$, the power against alternative $\beta \neq 0$ is approximately $\Phi(\delta - z_{\alpha/2}) + \Phi(-\delta - z_{\alpha/2})$, where $\delta = |\beta|/\text{SE}(\hat\beta)$ is the signal-to-noise ratio.
:::

```{r}
#| label: power-function
## Power for a two-sided t-test in OLS
power_ols <- function(effect, sigma_e, sd_x, n, alpha = 0.05) {
  se <- sigma_e / (sd_x * sqrt(n))
  delta <- abs(effect) / se
  z <- qnorm(1 - alpha / 2)
  pnorm(delta - z) + pnorm(-delta - z)
}
```

### Power curves

How does power vary with sample size and effect size?

```{r}
#| label: power-curves
#| fig-cap: "Power curves for a two-sided OLS t-test"
#| fig-width: 8
#| fig-height: 5
n_seq <- seq(20, 500, by = 5)
effects <- c(0.05, 0.10, 0.20, 0.40)

df_power <- do.call(rbind, lapply(effects, function(eff) {
  data.frame(
    n = n_seq,
    power = sapply(n_seq, function(nn) power_ols(eff, sigma_e = 1, sd_x = 1, nn)),
    effect = paste0("β = ", eff)
  )
}))

ggplot(df_power, aes(n, power, color = effect)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray50") +
  annotate("text", x = 480, y = 0.82, label = "80% power", color = "gray50") +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "gray70") +
  annotate("text", x = 480, y = 0.07, label = "α = 0.05", color = "gray70") +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Power curves for a two-sided OLS t-test",
       subtitle = "σ_e = 1, sd(X) = 1",
       x = "Sample size (n)", y = "Power", color = "True effect") +
  coord_cartesian(ylim = c(0, 1))
```

Small effects require enormous samples. With $\beta = 0.10$ and $\sigma_e / \text{sd}(X) = 1$, you need roughly $n = 800$ for 80% power.

::: {.callout-tip}
## The 80% Power Rule of Thumb
A study should have at least 80% power to detect the smallest effect of scientific interest. This requires $n \approx 16\sigma_e^2 / (\beta^2 \cdot \text{Var}(X))$ for a two-sided 5% test. Adding relevant controls reduces $\sigma_e$ and boosts power for free.
:::

### Simulation: observing power

Let's verify the formula with a Monte Carlo simulation:

```{r}
#| label: power-simulation
#| fig-width: 8
#| fig-height: 4
set.seed(12)
B <- 2000
true_beta <- 0.3
sigma_e <- 1
n_vals <- c(20, 50, 100, 200)

sim_power <- function(n, B) {
  rejections <- replicate(B, {
    x <- rnorm(n)
    y <- true_beta * x + rnorm(n, sd = sigma_e)
    pval <- summary(lm(y ~ x))$coefficients["x", "Pr(>|t|)"]
    pval < 0.05
  })
  mean(rejections)
}

results <- data.frame(
  n = n_vals,
  simulated = sapply(n_vals, sim_power, B = B),
  formula = sapply(n_vals, function(nn) power_ols(true_beta, sigma_e, 1, nn))
)
results
```

The simulated power matches the formula closely.

### Controls boost power

Adding a relevant control variable reduces $\sigma_e$ without reducing $\text{sd}(X)$, giving a free power boost:

```{r}
#| label: controls-power
set.seed(99)
B <- 2000
n <- 100
beta_x <- 0.2
beta_z <- 1.0  # strong control

## Without control
power_no_ctrl <- mean(replicate(B, {
  x <- rnorm(n); z <- rnorm(n)
  y <- beta_x * x + beta_z * z + rnorm(n)
  summary(lm(y ~ x))$coefficients["x", "Pr(>|t|)"] < 0.05
}))

## With control
power_with_ctrl <- mean(replicate(B, {
  x <- rnorm(n); z <- rnorm(n)
  y <- beta_x * x + beta_z * z + rnorm(n)
  summary(lm(y ~ x + z))$coefficients["x", "Pr(>|t|)"] < 0.05
}))

cat("Power without control:", round(power_no_ctrl, 3), "\n")
cat("Power with control:   ", round(power_with_ctrl, 3), "\n")
cat("Residual SD without z:", round(sqrt(beta_z^2 + 1), 2), "\n")
cat("Residual SD with z:   ", round(1, 2), "\n")
```

The control cuts the residual standard deviation roughly in half, dramatically increasing power. This is why pre-treatment covariates help in experiments even though they are unnecessary for unbiasedness.

### Joint tests have less power

Testing more restrictions simultaneously dilutes power. Fix the total non-centrality parameter and compare:

```{r}
#| label: joint-test-power
#| fig-width: 7
#| fig-height: 4
lambda_seq <- seq(0, 15, by = 0.1)

df_joint <- do.call(rbind, lapply(c(1, 2, 5, 10), function(q) {
  data.frame(
    lambda = lambda_seq,
    power = 1 - pchisq(qchisq(0.95, q), q, ncp = lambda_seq),
    q = paste0("q = ", q)
  )
}))

ggplot(df_joint, aes(lambda, power, color = q)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Power vs. non-centrality for joint tests",
       subtitle = "More restrictions require a larger signal to achieve the same power",
       x = expression(paste("Non-centrality parameter ", lambda)),
       y = "Power", color = "Restrictions")
```

A single-coefficient t-test is more powerful than a joint F-test that lumps in other restrictions. Test exactly what you care about.

## Statistical vs. economic significance

With a large enough sample, *any* nonzero coefficient becomes statistically significant. This is test consistency: $P[\text{Reject}] \to 1$ whenever $\beta \neq 0$.

```{r}
#| label: stat-vs-econ
#| fig-width: 8
#| fig-height: 4
set.seed(77)
true_beta <- 0.01   # tiny effect
n_vals <- c(50, 200, 1000, 5000, 20000)
B <- 1000

rejection_rate <- sapply(n_vals, function(nn) {
  mean(replicate(B, {
    x <- rnorm(nn)
    y <- true_beta * x + rnorm(nn)
    summary(lm(y ~ x))$coefficients["x", "Pr(>|t|)"] < 0.05
  }))
})

df_consist <- data.frame(n = n_vals, rejection_rate = rejection_rate)
ggplot(df_consist, aes(n, rejection_rate)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "coral") +
  scale_x_log10() +
  labs(title = "Test consistency: even β = 0.01 gets rejected eventually",
       subtitle = "With enough data, statistical significance is guaranteed for any nonzero effect",
       x = "Sample size (log scale)", y = "Rejection rate at α = 0.05")
```

A $p$-value tells you whether the data are inconsistent with $H_0$. It tells you nothing about whether the effect is large enough to matter. Always report the **magnitude** alongside the significance.


## Applied workflow: the salary example

Let's bring everything together with a thorough analysis of the professor salary data.

### Step 1: Fit and inspect

```{r}
#| label: salary-full
mod_full <- lm(salary ~ rank + discipline + yrs.since.phd + yrs.service + sex,
               data = Salaries)
coeftest(mod_full, vcov. = vcovHC(mod_full, type = "HC1"))
```

### Step 2: Key hypotheses

**Does rank matter?** (joint test, $q = 2$)

```{r}
#| label: salary-rank-test
linearHypothesis(mod_full, c("rankAssocProf = 0", "rankProf = 0"),
                 vcov. = vcovHC(mod_full, type = "HC1"))
```

**Does discipline matter?** (single restriction)

```{r}
#| label: salary-discipline-test
linearHypothesis(mod_full, "disciplineB = 0",
                 vcov. = vcovHC(mod_full, type = "HC1"))
```

**Are the two experience measures interchangeable?** ($\beta_{\text{phd}} = \beta_{\text{svc}}$)

```{r}
#| label: salary-experience-test
linearHypothesis(mod_full, "yrs.since.phd = yrs.service",
                 vcov. = vcovHC(mod_full, type = "HC1"))
```

### Step 3: Confidence intervals

```{r}
#| label: salary-ci
## Robust CIs
ci_robust <- coefci(mod_full, vcov. = vcovHC(mod_full, type = "HC1"))
ci_robust
```

### Step 4: Multiple testing adjustment

If we're testing all six regressors simultaneously:

```{r}
#| label: salary-multiple
pvals_all <- coeftest(mod_full, vcov. = vcovHC(mod_full, type = "HC1"))[-1, 4]
data.frame(
  raw_p = round(pvals_all, 4),
  holm_p = round(p.adjust(pvals_all, method = "holm"), 4),
  significant_raw = ifelse(pvals_all < 0.05, "*", ""),
  significant_holm = ifelse(p.adjust(pvals_all, method = "holm") < 0.05, "*", "")
)
```

### Step 5: Power assessment for the gender gap

Suppose the "true" gender gap is \$5,000 (about 5% of mean salary). Could we detect it?

```{r}
#| label: salary-power
sigma_e <- sigma(mod_full)
sd_sex <- sd(as.numeric(Salaries$sex == "Male"))
n <- nrow(Salaries)

power_sex <- power_ols(effect = 5000, sigma_e = sigma_e,
                       sd_x = sd_sex, n = n, alpha = 0.05)
cat("Estimated power to detect a $5,000 gender gap:", round(power_sex, 3), "\n")

## What sample size for 80% power?
n_needed <- seq(100, 5000, by = 10)
pow_curve <- sapply(n_needed, function(nn)
  power_ols(5000, sigma_e, sd_sex, nn, 0.05))
n80 <- n_needed[which(pow_curve >= 0.80)[1]]
cat("Sample size for 80% power:", n80, "\n")
```


## Practical advice (after Hansen)

A checklist for reporting regression results:

1. **Report standard errors, not just t-ratios.** Standard errors convey precision; t-statistics reduce everything to a binary yes/no.

2. **Report p-values, not asterisks.** The difference between $p = 0.049$ and $p = 0.051$ is not meaningful. Stars obscure this.

3. **Focus on substantive hypotheses.** Don't mechanically test every coefficient against zero. Test things that answer a scientific question.

4. **"Fail to reject" $\neq$ "accept."** Low power can explain non-rejection. Always ask: could we have detected a meaningful effect?

5. **Statistical significance $\neq$ economic significance.** A tiny but significant coefficient may be policy-irrelevant. A large but insignificant coefficient may reflect low power.

6. **Use robust inference by default.** There is rarely a good reason to report classical SEs when $n$ is not tiny. Use HC1/HC2 or the bootstrap.

7. **Account for multiple testing.** If you report the "most significant" result from many tests, apply Holm or Bonferroni corrections.


## Connection to GMM

The Wald test has a natural GMM interpretation. In GMM, testing $H_0: \mathbf{R}\boldsymbol{\beta} = \mathbf{r}$ is equivalent to testing whether the restricted moment conditions $E[\mathbf{Z}'(Y - \mathbf{X}\boldsymbol{\beta})] = \mathbf{0}$ with $\mathbf{R}\boldsymbol{\beta} = \mathbf{r}$ are compatible with the data. The J-test of overidentifying restrictions, which we'll meet in Chapter 11, generalizes the F-test to the setting where there are more moment conditions than parameters.

| OLS test | GMM counterpart |
|----------|-----------------|
| F-test (restricted vs. unrestricted) | Distance test (restricted vs. unrestricted GMM) |
| Wald test | Wald test (same formula, different $\hat{\mathbf{V}}$) |
| Score / LM test | Score test from restricted GMM |
| Overall F-statistic | J-test of overidentifying restrictions |
