---
title: "10. Constrained Least Squares"
subtitle: "Equality and inequality restrictions, and partial identification"
---

The previous chapter tested restrictions by comparing an unrestricted Wald statistic with a robust covariance matrix to its critical value. Some restricted-vs.-unrestricted machinery (the distance test, the LR-style criterion comparison) needs the *restricted* estimator: minimize SSE subject to the restriction. That is **constrained least squares** (CLS).

CLS is also useful in its own right:

- **Theory imposes structure.** Adding-up constraints (shares sum to 1), monotonicity (returns to schooling weakly increasing), sign restrictions (a price elasticity is non-positive) can all be encoded as restrictions on $\bm{\beta}$.
- **Partial identification.** When the data and minimal assumptions tell us only that $\bm{\beta}$ lies in a *set*, not a single point, inequality CLS lets us estimate that set and report the bounds.

This chapter covers:

1. Equality CLS — two ways to compute (reparameterize, or apply the closed-form formula)
2. The connection back to the Wald test (restricted SSE feeds the homoskedastic special case; the robust Wald is the right thing to report)
3. Inequality CLS — when the restriction is one-sided, via quadratic programming
4. A partial-identification application: bounds on the NSW employment ATE under missing data

```{r}
#| label: setup
#| message: false
library(sandwich)
library(lmtest)
library(estimatr)
library(car)
library(carData)
library(quadprog)   # QP for inequality CLS
library(lpSolve)    # LP for the bounds at the end
options(digits = 4)
data(Salaries)
```


## The setup

CLS minimizes the sum of squared errors subject to $q$ linear restrictions:
$$
\tilde{\bm{\beta}}_{\text{CLS}} \;=\; \arg\min_{\bm{\beta}} \; \sum_{i=1}^{n}(Y_i - \bm{X}_i'\bm{\beta})^2
\quad \text{s.t.} \quad \bm{R}'\bm{\beta} \;=\; \bm{r}
$$
where $\bm{R}$ is $k \times q$ and $\bm{r}$ is $q \times 1$. Three common forms:

| Restriction | $\bm{R}'$ | $\bm{r}$ |
|---|---|---|
| $\beta_3 = 0$ (exclusion) | $(0, 0, 1, 0)$ | $0$ |
| $\beta_2 = \beta_3$ (equality of two coefficients) | $(0, 1, -1, 0)$ | $0$ |
| $\beta_2 + \beta_3 = 1$ (adding-up) | $(0, 1, 1, 0)$ | $1$ |

The exclusion restriction is the trivial case --- "drop the regressor." The other two need either reparameterization or the closed-form formula below.


## Equality CLS by reparameterization

Use the `Salaries` data from `carData`. The model:
$$
\text{salary}_i = \alpha + \gamma_{\text{rank}} + \delta_{\text{discipline}} + \beta_2\,\text{yrs.since.phd}_i + \beta_3\,\text{yrs.service}_i + e_i.
$$
A natural restriction to test is $H_0: \beta_2 = \beta_3$ --- does an additional year since the PhD pay the same as an additional year of service at this institution? Becker--Mincer human-capital theory would predict the two should *differ* (firm-specific tenure should pay more), so the equality is something we test against, not something we believe a priori.

```{r}
#| label: cls-reparam
mod_U <- lm(salary ~ rank + discipline + yrs.since.phd + yrs.service,
            data = Salaries)

## Reparameterize: under H0, regress on the SUM of the two
mod_R <- lm(salary ~ rank + discipline + I(yrs.since.phd + yrs.service),
            data = Salaries)

cbind(unrestricted = c(coef(mod_U), NA),
      restricted   = c(coef(mod_R)[1:4], NA, NA, coef(mod_R)[5]))
```

In the restricted regression, the coefficient on `I(yrs.since.phd + yrs.service)` is the common $\tilde{\beta}_2 = \tilde{\beta}_3$. This is just OLS on a rebuilt design matrix --- no special machinery needed.

Note that the unrestricted estimates have *opposite signs* (+535 for `yrs.since.phd`, −477 for `yrs.service`): once `rank` is in the model, `yrs.service` measures extra years at the same rank without promotion, which signals lower productivity or weaker outside options and depresses salary. The two variables are nearly collinear ($\rho = 0.91$), so forcing them equal averages a strong positive against a strong negative --- the constrained common coefficient will land near zero, exactly the signal that the restriction is at odds with the data.


## Equality CLS via the closed-form formula

The general formula. Solving the constrained minimization with Lagrange multipliers gives
$$
\tilde{\bm{\beta}}_{\text{CLS}} \;=\; \hat{\bm{\beta}}_{\text{OLS}} - (\bm{X}'\bm{X})^{-1}\bm{R}\bigl[\bm{R}'(\bm{X}'\bm{X})^{-1}\bm{R}\bigr]^{-1}\bigl(\bm{R}'\hat{\bm{\beta}}_{\text{OLS}} - \bm{r}\bigr).
$$
Read it as: start from unrestricted OLS, subtract a correction proportional to *how far OLS violates the restriction*. If OLS already satisfies the restriction ($\bm{R}'\hat{\bm{\beta}}_{\text{OLS}} = \bm{r}$), the correction is zero.

```{r}
#| label: cls-formula
beta_OLS <- coef(mod_U)
X <- model.matrix(mod_U)
y <- Salaries$salary

## Build R for the restriction beta_phd = beta_service
R <- matrix(0, nrow = length(beta_OLS), ncol = 1)
R[which(names(beta_OLS) == "yrs.since.phd"), 1] <-  1
R[which(names(beta_OLS) == "yrs.service"),   1] <- -1
r <- 0

XX_inv <- solve(crossprod(X))
correction <- XX_inv %*% R %*%
  solve(t(R) %*% XX_inv %*% R) %*% (t(R) %*% beta_OLS - r)
beta_CLS <- as.vector(beta_OLS - correction)
names(beta_CLS) <- names(beta_OLS)
beta_CLS
```

The two `yrs.since.phd` and `yrs.service` entries should now be equal --- and equal to the coefficient on `I(yrs.since.phd + yrs.service)` from the reparameterized fit:

```{r}
#| label: cls-formula-check
common_value <- coef(mod_R)[["I(yrs.since.phd + yrs.service)"]]
c(formula_phd     = unname(beta_CLS["yrs.since.phd"]),
  formula_service = unname(beta_CLS["yrs.service"]),
  reparam_common  = common_value)
```

All three match. Two routes, same answer.


## Restricted SSE and the test

The headline test of $H_0: \beta_2 = \beta_3$ is the **robust Wald** built directly from the unrestricted estimator and a heteroskedasticity-robust covariance matrix. The restricted CLS estimator is not needed for this test --- but it is what makes the *classical F* well-defined as a homoskedastic special case.

```{r}
#| label: cls-robust-wald
## Headline: robust Wald (use this).
linearHypothesis(mod_U, "yrs.since.phd = yrs.service",
                 vcov. = vcovHC(mod_U, type = "HC1"))
```

Behind the scenes the classical F equals
$$
F \;=\; \frac{(\text{SSE}_R - \text{SSE}_U)/q}{\text{SSE}_U/(n - k)},
$$
a comparison of the *restricted* CLS sum of squares to the unrestricted OLS sum of squares. We can compute this from CLS directly and confirm `linearHypothesis()` returns the same thing under its homoskedastic default:

```{r}
#| label: cls-f-test
SSE_U <- sum(residuals(mod_U)^2)
SSE_R <- sum(residuals(mod_R)^2)
n     <- nrow(Salaries)
k     <- length(coef(mod_U))
q     <- 1

F_stat <- ((SSE_R - SSE_U) / q) / (SSE_U / (n - k))
p_val  <- 1 - pf(F_stat, q, n - k)
c(F = F_stat, p = p_val)
```

Under heteroskedasticity the classical $F$ has the wrong null distribution (it ignores the variation in $\text{Var}(e_i \mid X_i)$); the robust Wald above is what you should report. CLS earns its keep here mainly when you actually need the *restricted estimator itself* --- for inequality constraints (next sections), for the GMM distance test you saw in the IV chapter, or for imposing a theoretical structure as the headline regression.


## Inequality CLS

Equality CLS pins $\bm{\beta}$ to a $(k-q)$-dimensional subspace. **Inequality CLS** restricts $\bm{\beta}$ to a feasible region $\mathcal{B}$ defined by linear inequalities $\bm{R}'\bm{\beta} \le \bm{r}$ --- a polyhedron rather than a subspace. Examples:

- **Sign restrictions**: $\beta_j \ge 0$ (a variance, a non-negative elasticity, a share)
- **Box constraints**: $0 \le \beta_j \le 1$ (mixing weights, probabilities)
- **Monotonicity**: $\beta_1 \le \beta_2 \le \beta_3$ (ordered effects, e.g., dose-response that should be weakly increasing)

The strategy is two-step. Compute unrestricted OLS first. If $\hat{\bm{\beta}}_{\text{OLS}} \in \mathcal{B}$, no constraint binds and $\tilde{\bm{\beta}}_{\text{CLS}} = \hat{\bm{\beta}}_{\text{OLS}}$. Otherwise the constrained solution sits on the boundary of $\mathcal{B}$, where one or more constraints hold with equality. Conceptually you are *projecting* OLS onto the feasible region.

The general algorithm is a quadratic program: minimize $\frac{1}{2}\bm{\beta}'(\bm{X}'\bm{X})\bm{\beta} - (\bm{X}'\bm{y})'\bm{\beta}$ subject to the inequalities. In R, use `quadprog::solve.QP()`:

```{r}
#| label: cls-qp
## Suppose we want to estimate the salary regression but
## restrict the discipline coefficient to be NON-NEGATIVE
## (theoretical-economics premium, say).
mod_qp <- lm(salary ~ rank + discipline + yrs.since.phd + yrs.service,
             data = Salaries)
beta_OLS <- coef(mod_qp)
disc_idx <- which(names(beta_OLS) == "disciplineB")
cat("OLS coefficient on disciplineB:", round(beta_OLS[disc_idx], 1), "\n")

## QP setup
X <- model.matrix(mod_qp)
y <- Salaries$salary
Dmat <- crossprod(X)
dvec <- crossprod(X, y)

## Constraint: beta_disciplineB >= 0  (i.e. A' beta >= 0 with A picking that entry)
Amat <- matrix(0, nrow = ncol(X), ncol = 1)
Amat[disc_idx, 1] <- 1
bvec <- 0

qp <- solve.QP(Dmat = Dmat, dvec = dvec,
               Amat = Amat, bvec = bvec, meq = 0)
data.frame(coef = names(beta_OLS),
           OLS  = beta_OLS,
           CLS  = qp$solution,
           row.names = NULL)
```

OLS already gave a positive `disciplineB` coefficient, so the constraint doesn't bind --- inequality CLS returns the same answer. If the unrestricted estimate had been negative, `solve.QP` would return a solution with that coefficient pinned to exactly $0$ (the boundary) and the other coefficients adjusted accordingly.

A more interesting case: a negative-by-default coefficient with a sign restriction to test what the constrained estimate looks like. We force a `disciplineB` $\le 0$ constraint, which OLS *does* violate, and watch the projection happen:

```{r}
#| label: cls-qp-bind
## Reverse the constraint: require disciplineB <= 0 (unrealistic, illustrative)
Amat[disc_idx, 1] <- -1   # -1 * beta_disciplineB >= 0 means beta_disciplineB <= 0
bvec <- 0

qp_bind <- solve.QP(Dmat = Dmat, dvec = dvec,
                    Amat = Amat, bvec = bvec, meq = 0)
data.frame(coef = names(beta_OLS),
           OLS  = beta_OLS,
           CLS  = qp_bind$solution,
           row.names = NULL)
```

Now the constraint binds: the constrained `disciplineB` coefficient is exactly zero, and the other coefficients move to absorb the difference. This is what "projection onto the feasible region" looks like in practice.


## Partial identification: bounds on the NSW employment ATE

Inequality CLS becomes substantive when the *identified set* itself is a polyhedron. The setup: the National Supported Work Demonstration (Lalonde 1986; Dehejia & Wahba 1999) randomly assigned disadvantaged workers to a job-training program. Outcome: positive earnings in 1978.

```{r}
#| label: nsw-load
#| cache: true
u  <- "https://users.nber.org/~rdehejia/data/nswre74_"
cn <- c("treat","age","educ","black","hispan","married",
        "nodegree","re74","re75","re78")
nsw <- rbind(read.table(paste0(u, "treated.txt"), col.names = cn),
             read.table(paste0(u, "control.txt"), col.names = cn))
nsw$y <- as.numeric(nsw$re78 > 0)

y1 <- mean(nsw$y[nsw$treat == 1])
y0 <- mean(nsw$y[nsw$treat == 0])
naive_ate <- y1 - y0
c(y1 = y1, y0 = y0, naive_ATE = naive_ate)
```

The naive ATE on the experimental sample is about $0.11$ --- an 11 percentage-point employment effect. NSW's actual follow-up was nearly complete. The interesting question is what we could still pin down if a chunk of the sample had been *missing* at follow-up.

### Worst-case bounds

Suppose only $p = 70\%$ of subjects could be located in 1978. Within each arm, by the law of total probability,
$$E[Y \mid X = x] = p \cdot \bar{Y}_{\text{obs}}(x) + (1-p) \cdot \underbrace{E[Y \mid X = x, \text{missing}]}_{\text{unknown}}.$$
The unknown conditional mean lies in $[0, 1]$ (because $Y$ is binary), giving the *sharp* worst-case bounds:
$$p\,\bar{Y}_{\text{obs}}(x) \;\le\; E[Y \mid X{=}x] \;\le\; p\,\bar{Y}_{\text{obs}}(x) + (1-p).$$

```{r}
#| label: worst-case-bounds
p <- 0.7
lo_y1 <- p * y1;             hi_y1 <- p * y1 + (1 - p)
lo_y0 <- p * y0;             hi_y0 <- p * y0 + (1 - p)

ate_lo <- lo_y1 - hi_y0      # smallest ATE: arm 1 low, arm 0 high
ate_hi <- hi_y1 - lo_y0      # largest:      arm 1 high, arm 0 low

c(ATE_lo = ate_lo, ATE_hi = ate_hi, width = ate_hi - ate_lo)
```

Without any further assumption, the data + 30% missingness leaves the ATE pinned only to $[-0.22, 0.38]$ --- wide enough that we cannot even sign the effect.

### Tighter bounds via monotone selection

A mild (and often credible) assumption: those *lost* to follow-up are weakly *less* likely to be employed than those we observed in the same arm. Formally, $E[Y \mid X{=}x, \text{missing}] \le \bar{Y}_{\text{obs}}(x)$. This tightens the upper bound (the missing group can't contribute more than $\bar{Y}_{\text{obs}}$), leaving the lower bound unchanged:

```{r}
#| label: mts-bounds
hi_y1_mts <- y1   # MTS: replace 1 with Y_obs
hi_y0_mts <- y0

ate_lo_mts <- lo_y1 - hi_y0_mts
ate_hi_mts <- hi_y1_mts - lo_y0

c(ATE_lo = ate_lo_mts, ATE_hi = ate_hi_mts, width = ate_hi_mts - ate_lo_mts)
```

The width drops from $0.60$ to $0.42$. The interval still crosses zero --- an 11 pp underlying effect is too small for one mild assumption to settle the sign at 30% missing --- but the bounds *are* informative now.

### The identified set as inequality CLS via `lpSolve`

Both bounds above came from closed-form arithmetic. With more constraints --- multiple monotonicity assumptions, or a parameter that's not just $\mu_1 - \mu_0$ --- the closed-form gets unwieldy. Phrase it instead as a **linear program** over the feasible region.

Reparameterize from the regression coefficients to the conditional means $\bm{\mu} = (\mu_0, \mu_1)$ where $\mu_x = E[Y \mid X{=}x]$. The MTS bounds become four inequalities:
$$
0.452 \le \mu_0 \le 0.646, \qquad 0.530 \le \mu_1 \le 0.757.
$$
The ATE is $\mu_1 - \mu_0 = \bm{c}'\bm{\mu}$ with $\bm{c} = (-1, 1)'$.

```{r}
#| label: lp-bounds
A   <- rbind(c(1,0), c(1,0), c(0,1), c(0,1))
dir <- c(">=", "<=", ">=", "<=")
rhs <- c(lo_y0, hi_y0_mts, lo_y1, hi_y1_mts)
c_ate <- c(-1, 1)

c(ATE_lo = lp("min", c_ate, A, dir, rhs)$objval,
  ATE_hi = lp("max", c_ate, A, dir, rhs)$objval)
```

Same answer (`ATE_lo` ≈ −0.12, `ATE_hi` ≈ 0.30), but now the recipe extends to any number of constraints. Adding another assumption is just another row in `A`, `dir`, `rhs`.


## What CLS is for, in one sentence

The Wald test handles most "is this restriction true?" questions without ever computing the restricted estimator. CLS earns its place when you need the *restricted estimator itself*: imposing a theoretical structure as the headline regression (equality form), or estimating an *identified set* when the data alone do not pin $\bm{\beta}$ down (inequality form). Both reduce to the same idea: minimize SSE on a restricted set of parameter values.
