---
title: "13. Panel Data"
subtitle: "Pooled, between, and within estimators; difference-in-differences; dynamic panels"
---

This chapter covers panel data methods: pooled OLS, the between and within (fixed effects) estimators, difference-in-differences, clustering, and dynamic panel GMM. We use the `plm` package for panel estimation, `panelView` for visualization, and `fixest` for fast fixed effects with flexible standard errors.

**Questions this chapter answers:**

1. How do `panelView` and `ggplot` help visualize treatment patterns, missingness, and outcome dynamics?
2. What are the pooled, between, and within estimators — and when is each consistent?
3. How does difference-in-differences identify causal effects under parallel trends?
4. How do Arellano-Bond and system GMM handle dynamic panels with lagged dependent variables?

```{r}
#| label: setup
#| message: false
library(ggplot2)
library(plm)
library(panelView)
library(fixest)
library(sandwich)
library(lmtest)
library(estimatr)
```

## Visualizing panel structure with panelView

Before estimating anything, it pays to *look* at your panel. The `panelView` package visualizes treatment patterns, missingness, and outcome dynamics.

```{r}
#| label: panelview-datasets
# Load the built-in datasets
data(panelView)
```

### Treatment status plots

The `turnout` dataset tracks U.S. state voter turnout and election-day registration (EDR) policy adoption---a classic staggered treatment design (see the [TWFE bias discussion](ch13-fixed-effects.qmd#sec-twfe-bias) for the problem under staggered adoption).

```{r}
#| label: panelview-treat
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State",
          main = "Election-day registration adoption by state")
```

Sorting by treatment timing reveals the staggered adoption pattern:

```{r}
#| label: panelview-timing
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          by.timing = TRUE,
          xlab = "Year", ylab = "State",
          main = "EDR adoption sorted by timing")
```

### Missing data plots

The `capacity` dataset (country-level state capacity) has substantial missingness. `panelView` reveals the pattern:

```{r}
#| label: panelview-missing
panelview(Capacity ~ demo + lngdp + lnpop,
          data = capacity, index = c("ccode", "year"),
          type = "miss", axis.lab = "off",
          main = "Missing data in state capacity panel")
```

### Outcome dynamics

We can also plot the outcome variable over time, colored by treatment status:

```{r}
#| label: panelview-outcome
panelview(turnout ~ policy_edr,
          data = turnout, index = c("abb", "year"),
          type = "outcome",
          main = "Voter turnout by EDR status",
          ylab = "Turnout (%)", xlab = "Year")
```

## Panel data basics: the Produc dataset

We use the `Produc` dataset from the `plm` package: 48 U.S. states observed from 1970--1986, with gross state product, public capital, private capital, employment, and unemployment.

```{r}
#| label: produc-setup
data("Produc", package = "plm")
cat("States:", length(unique(Produc$state)),
    " Years:", min(Produc$year), "-", max(Produc$year),
    " Obs:", nrow(Produc), "\n")
head(Produc[, c("state", "year", "gsp", "pcap", "emp", "unemp")])
```

We model log gross state product as a function of log public capital, log private capital, log employment, and unemployment rate:

$$\ln(\text{gsp})_{it} = \beta_1 \ln(\text{pcap})_{it} + \beta_2 \ln(\text{pc})_{it} + \beta_3 \ln(\text{emp})_{it} + \beta_4 \text{unemp}_{it} + \alpha_i + \varepsilon_{it}$$

The question is whether $\alpha_i$ (the state-specific intercept) is correlated with the regressors.

### Visualizing the panel

```{r}
#| label: produc-plot
# Log GSP over time by state
ggplot(Produc, aes(x = year, y = log(gsp), group = state)) +
  geom_line(alpha = 0.3) +
  labs(title = "Log gross state product over time",
       x = "Year", y = "log(GSP)") +
  theme_minimal()
```

## Pooled OLS

Pooled OLS ignores the panel structure entirely---it treats all $NT$ observations as independent:

```{r}
#| label: pooled-ols
pooled <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
              data = Produc, index = c("state", "year"),
              model = "pooling")
summary(pooled)
```

Pooled OLS is consistent only if the unobserved state effects $\alpha_i$ are uncorrelated with all regressors. If wealthier states invest more in public capital (likely), pooled OLS is biased.

## The between estimator

The between estimator uses only cross-sectional variation---it regresses group means $\bar{y}_i$ on $\bar{x}_i$:

```{r}
#| label: between
between_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
                   data = Produc, index = c("state", "year"),
                   model = "between")
summary(between_fit)
```

The between estimator discards all within-state time variation. It is consistent only under the random effects assumption ($\alpha_i \perp X_{it}$). When that assumption fails, the between estimator is biased.

## The within (fixed effects) estimator {#sec-within}

The within estimator removes $\alpha_i$ by demeaning each variable within each state:

$$(\ln\text{gsp}_{it} - \overline{\ln\text{gsp}}_i) = \beta'(x_{it} - \bar{x}_i) + (\varepsilon_{it} - \bar{\varepsilon}_i)$$ {#eq-within}

::: {#def-within}
## Within (Fixed Effects) Estimator
The within estimator demeans all variables within each unit: $(y_{it} - \bar{y}_i) = \beta'(x_{it} - \bar{x}_i) + (\varepsilon_{it} - \bar\varepsilon_i)$. It is consistent when $\mathbb{E}[\varepsilon_{it}|X_{i1}, \ldots, X_{iT}, \alpha_i] = 0$, even if $\alpha_i$ is correlated with $X_{it}$.
:::

::: {.callout-note}
## Within Eliminates Time-Invariant Confounders
The within estimator removes all time-invariant unit characteristics $\alpha_i$ — both observed and unobserved. This eliminates omitted variable bias from time-invariant confounders, but it also prevents estimating effects of time-invariant regressors (e.g., region, ethnicity). Use CRE/Mundlak (Chapter 13) to recover these.
:::

```{r}
#| label: within
within_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
                  data = Produc, index = c("state", "year"),
                  model = "within")
summary(within_fit)
```

### Fixed effects by hand

By the [FWL theorem](ch04-sensitivity.qmd#thm-fwl), the within estimator is equivalent to including dummy variables for each state (LSDV), which is equivalent to the Frisch--Waugh--Lovell projection:

```{r}
#| label: within-by-hand
# LSDV approach: include state dummies
lsdv <- lm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp + factor(state),
           data = Produc)

# Compare slope coefficients
cat("plm within:  ", round(coef(within_fit), 4), "\n")
cat("LSDV slopes: ", round(coef(lsdv)[2:5], 4), "\n")
```

### Manual demeaning

```{r}
#| label: demean-by-hand
# Demean each variable within state
demean <- function(x, group) x - ave(x, group)

Produc$lgsp_dm  <- demean(log(Produc$gsp), Produc$state)
Produc$lpcap_dm <- demean(log(Produc$pcap), Produc$state)
Produc$lpc_dm   <- demean(log(Produc$pc), Produc$state)
Produc$lemp_dm  <- demean(log(Produc$emp), Produc$state)
Produc$unemp_dm <- demean(Produc$unemp, Produc$state)

dm_fit <- lm(lgsp_dm ~ lpcap_dm + lpc_dm + lemp_dm + unemp_dm - 1, data = Produc)
cat("Manual demeaning: ", round(coef(dm_fit), 4), "\n")
cat("plm within:       ", round(coef(within_fit), 4), "\n")
```

### Comparing estimators

```{r}
#| label: compare-estimators
coef_table <- data.frame(
  Variable = names(coef(within_fit)),
  Pooled = round(coef(pooled)[-1], 4),
  Between = round(coef(between_fit)[-1], 4),
  Within = round(coef(within_fit), 4)
)
rownames(coef_table) <- NULL
coef_table
```

The pooled and between estimators give a large positive coefficient on public capital (`pcap`). The within estimator---which controls for unobserved state characteristics---gives a much smaller (or negative) coefficient. This is the classic omitted variable bias story: states with high unobserved productivity invest in more public capital *and* have higher GSP.

### Variance decomposition

The total variation in a panel decomposes into between-group and within-group components. The pooled estimator is a matrix-weighted average of the between and within estimators:

```{r}
#| label: variance-decomp
# Between variation
x_bar <- aggregate(log(pcap) ~ state, data = Produc, mean)$`log(pcap)`
between_var <- var(x_bar)

# Within variation
within_var <- var(demean(log(Produc$pcap), Produc$state))

# Total
total_var <- var(log(Produc$pcap))

cat("Between variance:", round(between_var, 4), "\n")
cat("Within variance: ", round(within_var, 4), "\n")
cat("Total variance:  ", round(total_var, 4), "\n")
cat("Between share:   ", round(between_var / total_var * 100, 1), "%\n")
```

When the between share is large (as here), pooled OLS is heavily influenced by cross-sectional differences between states, which may be confounded by omitted variables.

## Clustering and robust standard errors

Standard errors from pooled OLS or even fixed effects may be wrong if errors are correlated within states (serial correlation) or across states (spatial correlation).

### Cluster-robust standard errors

```{r}
#| label: cluster-se
# Default plm standard errors (assume iid errors)
cat("Default within SE:\n")
round(coef(summary(within_fit))[, "Std. Error"], 4)
```

```{r}
#| label: cluster-robust
# Cluster-robust SE (Arellano, 1987)
cat("\nCluster-robust SE (Arellano):\n")
round(sqrt(diag(vcovHC(within_fit, method = "arellano", type = "HC1"))), 4)
```

```{r}
#| label: newey-west-panel
# Newey-West SE for serial correlation
cat("\nNewey-West SE (1 lag):\n")
round(sqrt(diag(vcovNW(within_fit, maxlag = 1))), 4)
```

::: {.callout-warning}
## Always Cluster Standard Errors in Panel Data
Standard OLS errors assume independence across observations. In panel data, errors are typically correlated within units (serial correlation). Failing to cluster inflates test statistics by the Moulton factor $\sqrt{1 + (T-1)\rho}$, which can be substantial.
:::

Using `fixest::feols()` provides the same estimates with more flexible standard error options:

```{r}
#| label: fixest-fe
fe_fixest <- feols(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp | state,
                   data = Produc, panel.id = ~state + year)

# Various SE types
summary(fe_fixest, vcov = "iid")
```

```{r}
#| label: fixest-cluster
summary(fe_fixest, vcov = ~state)  # cluster by state
```

```{r}
#| label: fixest-driscoll-kraay
# Driscoll-Kraay SE (cross-sectional + serial dependence)
summary(fe_fixest, vcov = "DK")
```

### The Moulton factor

Why do clustered SEs matter? When errors are correlated within clusters, standard SEs are too small by a factor that grows with cluster size and intra-cluster correlation $\rho$:

$$\text{Moulton factor} \approx \sqrt{1 + (n_g - 1)\rho}$$

```{r}
#| label: moulton
# Estimate intra-state error correlation from within-model residuals
e_within <- residuals(within_fit)
pdat <- pdata.frame(Produc, index = c("state", "year"))
state_ids <- as.numeric(pdat$state)

# Compute average within-state residual correlation
states <- unique(state_ids)
cors <- numeric(length(states))
for (j in seq_along(states)) {
  e_j <- e_within[state_ids == states[j]]
  if (length(e_j) > 1) {
    pair_cors <- cor(e_j[-length(e_j)], e_j[-1])
    cors[j] <- pair_cors
  }
}
rho <- mean(cors, na.rm = TRUE)
T_per_state <- nrow(Produc) / length(unique(Produc$state))
moulton <- sqrt(1 + (T_per_state - 1) * rho)
cat("Intra-state correlation (rho):", round(rho, 3), "\n")
cat("Average T per state:", T_per_state, "\n")
cat("Moulton factor:", round(moulton, 2), "\n")
```

Standard errors should be inflated by roughly this factor when clustering is ignored.

## Difference-in-differences {#sec-did}

Difference-in-differences (DiD) is a panel method for estimating causal effects when treatment is assigned to some units at some time. The canonical $2 \times 2$ DiD model is:

$$Y_{it} = \alpha + \beta \cdot \text{Group}_i + \gamma \cdot \text{Post}_t + \delta \cdot (\text{Group}_i \times \text{Post}_t) + \varepsilon_{it}$$ {#eq-did}

::: {#thm-did}
## Difference-in-Differences
The DiD estimator $\hat\delta = (\bar{Y}_{1,\text{post}} - \bar{Y}_{1,\text{pre}}) - (\bar{Y}_{0,\text{post}} - \bar{Y}_{0,\text{pre}})$ identifies the ATT under the **parallel trends** assumption: absent treatment, treated and control groups would have followed the same trajectory.
:::

where $\delta$ is the treatment effect, identified under the **parallel trends** assumption.

### Simulated DiD example

```{r}
#| label: did-sim
set.seed(42)
N <- 100  # units
T_periods <- 10
treat_time <- 6  # treatment starts at t = 6
treated_units <- 1:(N/2)

did_data <- expand.grid(unit = 1:N, time = 1:T_periods)
did_data$treated <- as.integer(did_data$unit %in% treated_units)
did_data$post <- as.integer(did_data$time >= treat_time)

# Generate outcomes with parallel pre-trends
alpha_i <- rep(rnorm(N, sd = 2), each = T_periods)  # unit effects
gamma_t <- rep(0.3 * (1:T_periods), times = N)       # common trend
delta_true <- 1.5

did_data$y <- 5 + alpha_i + gamma_t +
  delta_true * did_data$treated * did_data$post +
  rnorm(N * T_periods, sd = 1)
```

```{r}
#| label: did-plot
#| fig-cap: "Difference-in-differences: group means over time with treatment onset"
# Group means over time
group_means <- aggregate(y ~ time + treated, data = did_data, mean)
group_means$group <- factor(group_means$treated, labels = c("Control", "Treated"))

ggplot(group_means, aes(x = time, y = y, color = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = treat_time - 0.5, linetype = "dashed") +
  annotate("text", x = treat_time - 0.3, y = max(group_means$y),
           label = "Treatment", hjust = 1, size = 3) +
  labs(title = "Difference-in-differences",
       x = "Time", y = "Outcome", color = "Group") +
  theme_minimal()
```

### DiD estimation

```{r}
#| label: did-estimate
# Method 1: OLS with interaction
did_ols <- lm(y ~ treated * post, data = did_data)
cat("DiD coefficient (OLS interaction):", round(coef(did_ols)["treated:post"], 3),
    " (true:", delta_true, ")\n")
```

```{r}
#| label: did-twfe
# Method 2: Two-way fixed effects (unit + time FE)
did_twfe <- feols(y ~ treated:post | unit + time, data = did_data)
cat("DiD coefficient (TWFE):", round(coef(did_twfe), 3), "\n")
```

```{r}
#| label: did-cluster
# With cluster-robust SEs (cluster on unit)
summary(did_twfe, vcov = ~unit)
```

### Visualizing the DiD with panelView

```{r}
#| label: did-panelview
# Treatment pattern
did_data$D <- did_data$treated * did_data$post
panelview(y ~ D, data = did_data, index = c("unit", "time"),
          type = "treat",
          main = "Treatment pattern in DiD design")
```

## Random effects

The random effects (RE) estimator assumes $\alpha_i \perp X_{it}$ and estimates a GLS model with the composite error $\nu_{it} = \alpha_i + \varepsilon_{it}$. The RE estimator is a weighted average of the between and within estimators:

$$\hat{\beta}_{RE} = \lambda \hat{\beta}_{W} + (1 - \lambda) \hat{\beta}_{B}$$

where $\lambda$ depends on the ratio of within to between variance.

```{r}
#| label: random-effects
re_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
              data = Produc, index = c("state", "year"),
              model = "random")
summary(re_fit)
```

### Hausman test: FE vs RE

The Hausman test compares FE and RE. Under $H_0$ (RE is consistent), both estimators are consistent but RE is efficient. Under $H_1$, only FE is consistent.

```{r}
#| label: hausman
phtest(within_fit, re_fit)
```

A small p-value means we reject the RE assumption---the fixed effects are correlated with the regressors, so FE is preferred.

### Mundlak's correlated random effects

Mundlak (1978) showed that we can relax the RE assumption by including the group means $\bar{x}_i$ as additional regressors. This "correlated random effects" (CRE) estimator gives the same slope coefficients as FE while also estimating effects of time-invariant variables:

```{r}
#| label: mundlak
# Compute state means of the regressors
Produc$lpcap_mean <- ave(log(Produc$pcap), Produc$state)
Produc$lpc_mean   <- ave(log(Produc$pc), Produc$state)
Produc$lemp_mean  <- ave(log(Produc$emp), Produc$state)
Produc$unemp_mean <- ave(Produc$unemp, Produc$state)

# CRE = RE + group means
cre_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp +
                 lpcap_mean + lpc_mean + lemp_mean + unemp_mean,
               data = Produc, index = c("state", "year"),
               model = "random")

# Compare CRE slopes to FE slopes
cat("CRE slopes: ", round(coef(cre_fit)[2:5], 4), "\n")
cat("FE slopes:  ", round(coef(within_fit), 4), "\n")
```

The slope coefficients on the time-varying regressors match the FE estimates, confirming the Mundlak equivalence. The coefficients on the group means capture the between-group relationships that FE removes.

## Dynamic panels and Arellano--Bond GMM

When the model includes a lagged dependent variable, fixed effects estimation is inconsistent (Nickell bias, $O(1/T)$). Arellano and Bond (1991) proposed first-differencing the equation and using lagged levels as instruments:

$$\Delta y_{it} = \rho \Delta y_{it-1} + \Delta x_{it}'\beta + \Delta \varepsilon_{it}$$

Since $\text{Cov}(y_{is}, \Delta\varepsilon_{it}) = 0$ for $s \leq t-2$, lagged levels are valid instruments. This creates a growing set of moment conditions---naturally suited to GMM.

### The EmplUK dataset

We use the `EmplUK` dataset: 140 UK firms observed over 1976--1984, with employment, wages, capital, and output.

```{r}
#| label: empluk-setup
data("EmplUK", package = "plm")
cat("Firms:", length(unique(EmplUK$firm)),
    " Years:", min(EmplUK$year), "-", max(EmplUK$year),
    " Obs:", nrow(EmplUK), "\n")
head(EmplUK)
```

```{r}
#| label: empluk-panelview
# Visualize the panel structure
panelview(emp ~ wage, data = EmplUK,
          index = c("firm", "year"),
          type = "miss", axis.lab = "off",
          main = "EmplUK panel balance")
```

### Nickell bias: why FE fails with lagged DV

```{r}
#| label: nickell-sim
set.seed(123)
N_nick <- 100; T_nick <- 8
rho_true <- 0.5

# Simulate dynamic panel
nick_data <- data.frame()
for (i in 1:N_nick) {
  alpha_i <- rnorm(1, sd = 1)
  y <- numeric(T_nick)
  y[1] <- alpha_i / (1 - rho_true) + rnorm(1)
  for (t in 2:T_nick) {
    y[t] <- rho_true * y[t-1] + alpha_i + rnorm(1)
  }
  nick_data <- rbind(nick_data,
    data.frame(id = i, time = 1:T_nick, y = y))
}

# FE estimate of rho (biased downward)
nick_data$y_lag <- ave(nick_data$y, nick_data$id,
                       FUN = function(x) c(NA, x[-length(x)]))
fe_nick <- plm(y ~ y_lag, data = nick_data, index = c("id", "time"),
               model = "within", na.action = na.omit)
cat("True rho:", rho_true, "\n")
cat("FE estimate:", round(coef(fe_nick), 4), " (Nickell bias)\n")
cat("Expected bias ~ -1/T =", round(-1/T_nick, 4), "\n")
```

The FE estimator is biased downward by approximately $-1/T$ when $T$ is small.

::: {#def-nickell-bias}
## Nickell Bias
In a dynamic panel $y_{it} = \rho y_{it-1} + \alpha_i + \varepsilon_{it}$, fixed effects estimation is inconsistent with bias $\approx -(\rho + 1)/(T - 1)$. This arises because demeaning creates mechanical correlation between $\tilde{y}_{it-1}$ and $\tilde\varepsilon_{it}$. The bias is severe when $T$ is small.
:::

### Arellano--Bond difference GMM

The `pgmm()` function in `plm` implements Arellano--Bond (difference GMM) and Blundell--Bond (system GMM).

The formula syntax uses `|` to separate regressors from GMM instruments:

```
y ~ regressors | GMM_instruments | standard_instruments
```

```{r}
#| label: arellano-bond
# Arellano-Bond: Table 4, Column (b) from Arellano & Bond (1991)
ab_fit <- pgmm(
  log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1)
            + log(capital) + lag(log(output), 0:1)
            | lag(log(emp), 2:99),
  data = EmplUK,
  effect = "twoways",
  model = "twosteps",
  transformation = "d"  # difference GMM
)

summary(ab_fit, robust = TRUE)
```

### Diagnostic tests

Two diagnostics are essential for dynamic panel GMM:

1. **Sargan/Hansen J-test**: Are the overidentifying restrictions satisfied?
2. **AR(2) test**: Is there second-order serial correlation in the differenced residuals? (AR(1) is expected due to differencing; AR(2) would invalidate lag-2 instruments.)

```{r}
#| label: ab-diagnostics
# Sargan test of overidentifying restrictions
sargan(ab_fit)
```

```{r}
#| label: ab-ar-tests
# Serial correlation tests
mtest(ab_fit, order = 1)  # AR(1): expected to reject
mtest(ab_fit, order = 2)  # AR(2): should not reject
```

AR(1) in the differenced residuals is expected and mechanical. The key test is AR(2): failing to reject means our instruments (lags 2+) are valid.

### Blundell--Bond system GMM

When the dependent variable is highly persistent ($\rho$ close to 1), lagged levels become weak instruments for first differences. Blundell and Bond (1998) augment the moment conditions with level equations, using lagged *differences* as instruments:

```{r}
#| label: system-gmm
# System GMM: use transformation = "ld" (level + difference)
bb_fit <- pgmm(
  log(emp) ~ lag(log(emp), 1) + lag(log(wage), 0:1)
            + lag(log(capital), 0:1)
            | lag(log(emp), 2:99) + lag(log(wage), 2:99)
              + lag(log(capital), 2:99),
  data = EmplUK,
  effect = "twoways",
  model = "twosteps",
  transformation = "ld"  # system GMM
)

summary(bb_fit, robust = TRUE)
```

```{r}
#| label: bb-diagnostics
sargan(bb_fit)
mtest(bb_fit, order = 2)
```

### Instrument proliferation

A practical concern with `pgmm()` is that the number of GMM instruments grows quadratically with $T$. With $T = 9$ periods, using `lag(y, 2:99)` generates up to $(T-1)(T-2)/2 = 28$ instruments from a single variable. Too many instruments can overfit the endogenous variable.

The `collapse = TRUE` option reduces the instrument count:

```{r}
#| label: collapse-instruments
# Collapsed instrument matrix
ab_collapse <- pgmm(
  log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1)
            + log(capital) + lag(log(output), 0:1)
            | lag(log(emp), 2:99),
  data = EmplUK,
  effect = "twoways",
  model = "twosteps",
  transformation = "d",
  collapse = TRUE
)

summary(ab_collapse, robust = TRUE)
```

```{r}
#| label: collapse-comparison
# Compare coefficient on lag(log(emp), 1) across specifications
cat("Full instruments:     ", round(coef(ab_fit)["lag(log(emp), 1:2)1"], 4), "\n")
cat("Collapsed instruments:", round(coef(ab_collapse)["lag(log(emp), 1:2)1"], 4), "\n")
```

## Panel estimators as moment conditions

Every panel estimator we've covered is a GMM estimator with specific moment conditions:

| Estimator | Moment condition | Instruments |
|-----------|-----------------|-------------|
| Pooled OLS | $\mathbb{E}[X_{it} \varepsilon_{it}] = 0$ | $X_{it}$ |
| Within (FE) | $\mathbb{E}[\tilde{X}_{it} \tilde{\varepsilon}_{it}] = 0$ | Demeaned $\tilde{X}_{it}$ |
| Between | $\mathbb{E}[\bar{X}_i \bar{\varepsilon}_i] = 0$ | Group means $\bar{X}_i$ |
| Random effects | $\mathbb{E}[X_{it} \nu_{it}] = 0$, $\mathbb{E}[\bar{X}_i \nu_{it}] = 0$ | Both $X_{it}$ and $\bar{X}_i$ |
| Arellano--Bond | $\mathbb{E}[y_{is} \Delta\varepsilon_{it}] = 0,\ s \leq t-2$ | Lagged levels |
| Blundell--Bond | Above + $\mathbb{E}[\Delta y_{is} \varepsilon_{it}] = 0,\ s \leq t-1$ | + Lagged differences |

The progression from OLS to system GMM mirrors the course arc: each step adds moment conditions and addresses a new identification challenge.

## Choosing a panel estimator

```{r}
#| label: decision-tree
#| echo: false
cat("
Panel Estimator Decision Guide:
================================
1. Is alpha_i correlated with X_it?
   - No  --> Random Effects (more efficient)
   - Yes --> Fixed Effects
   - Unsure --> Run Hausman test

2. Do you have a lagged dependent variable?
   - No  --> Standard FE is fine
   - Yes --> FE is biased (Nickell bias)
           --> Use Arellano-Bond or System GMM

3. Is the DV highly persistent (rho near 1)?
   - No  --> Arellano-Bond (difference GMM)
   - Yes --> Blundell-Bond (system GMM)

4. Diagnostics:
   - Always report cluster-robust SEs
   - For GMM: check Sargan test + AR(2) test
   - Watch for instrument proliferation
")
```

## Summary

- **panelView** visualizes treatment adoption, missingness, and outcome dynamics before estimation.
- **Pooled OLS** ignores unit effects; **between** uses only cross-sectional variation; **within (FE)** uses only time variation within units.
- The within estimator removes time-invariant confounders but cannot estimate effects of time-invariant regressors.
- **Cluster-robust SEs** (Arellano, Newey--West, Driscoll--Kraay) account for within-cluster error dependence. The Moulton factor quantifies the standard error inflation from ignoring clustering.
- **Mundlak's CRE** gives FE slopes while retaining the ability to estimate effects of time-invariant variables.
- **Hausman test** compares FE and RE; rejection means $\alpha_i$ is correlated with $X_{it}$.
- **Arellano--Bond** and **Blundell--Bond** use GMM for dynamic panels with lagged dependent variables, overcoming Nickell bias.
- **Diagnostic tests**: Sargan/Hansen J-test for overidentifying restrictions; AR(2) test for instrument validity.
- All panel estimators are special cases of GMM with different moment conditions.
