---
title: "14. Fixed Effects and Modern DiD"
subtitle: "Random effects, Hausman test, CRE, dynamic panels, and new DiD methods"
---

This chapter covers the fixed vs. random effects choice, correlated random effects (Mundlak), dynamic panel GMM, and modern difference-in-differences methods that address the problems with two-way fixed effects under staggered treatment adoption. We use `plm` for panel estimation, `did` for Callaway--Sant'Anna group-time ATTs, and `fect` for counterfactual imputation estimators.

**Questions this chapter answers:**

1. When should you use fixed effects vs. random effects, and what does the Hausman test tell you?
2. How does the TWFE estimator fail under staggered treatment with heterogeneous effects?
3. What modern alternatives (Callaway-Sant'Anna, fect, matrix completion) fix the negative weighting problem?

```{r}
#| label: setup
#| message: false
library(ggplot2)
library(plm)
library(fixest)
library(sandwich)
library(lmtest)
library(did)
library(fect)
```

## Random effects as GLS

In Chapter 12 we estimated fixed effects by demeaning within each unit. The **random effects** (RE) estimator instead treats $\alpha_i$ as part of a composite error $\nu_{it} = \alpha_i + \varepsilon_{it}$ and applies GLS to exploit the known correlation structure:

$$\text{Var}(\nu_i) = \sigma_\varepsilon^2 I_T + \sigma_\alpha^2 \iota\iota' = \Omega$$ {#eq-re-variance}

```{r}
#| label: produc-setup
data("Produc", package = "plm")
pdata <- pdata.frame(Produc, index = c("state", "year"))
```

```{r}
#| label: re-estimation
# Fixed effects
fe_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
              data = pdata, model = "within")

# Random effects (GLS with error components)
re_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
              data = pdata, model = "random")

# Compare slope coefficients
coef_comp <- data.frame(
  Variable = names(coef(fe_fit)),
  FE = round(coef(fe_fit), 4),
  RE = round(coef(re_fit)[-1], 4)
)
rownames(coef_comp) <- NULL
coef_comp
```

RE is more efficient than FE when the random effects assumption ($\alpha_i \perp X_{it}$) holds, because it uses both within-unit and between-unit variation. But if $\alpha_i$ is correlated with the regressors, RE is inconsistent.

### Variance components

The RE estimator decomposes total variance into unit-level ($\sigma_\alpha^2$) and idiosyncratic ($\sigma_\varepsilon^2$) components:

```{r}
#| label: variance-components
# Extract variance components from plm
ercomp <- ercomp(re_fit)
cat("sigma_alpha (unit):", round(sqrt(ercomp$sigma2["id"]), 4), "\n")
cat("sigma_eps (idios):", round(sqrt(ercomp$sigma2["idios"]), 4), "\n")
cat("theta (shrinkage):", round(ercomp$theta, 4), "\n")
```

The shrinkage parameter $\theta$ controls how much RE pulls toward the within estimator versus the between estimator. When $\theta$ is close to 1, RE is close to FE; when $\theta$ is close to 0, RE is close to pooled OLS.

### Partial pooling: RE as a weighted average

```{r}
#| label: partial-pooling
# Between estimator
be_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp,
              data = pdata, model = "between")

cat("Between: ", round(coef(be_fit)["log(pcap)"], 4), "\n")
cat("Within:  ", round(coef(fe_fit)["log(pcap)"], 4), "\n")
cat("RE:      ", round(coef(re_fit)["log(pcap)"], 4), "\n")
cat("(RE is between the within and between estimates)\n")
```

## The Hausman test

The Hausman test compares FE (consistent under both $H_0$ and $H_1$) with RE (efficient under $H_0$ but inconsistent under $H_1$):

$$H = (\hat\beta_{FE} - \hat\beta_{RE})' [\text{Var}(\hat\beta_{FE}) - \text{Var}(\hat\beta_{RE})]^{-1} (\hat\beta_{FE} - \hat\beta_{RE}) \xrightarrow{d} \chi^2_K$$ {#eq-hausman}

::: {#thm-hausman}
## Hausman Test
The Hausman statistic $H = (\hat\beta_{FE} - \hat\beta_{RE})'[\text{Var}(\hat\beta_{FE}) - \text{Var}(\hat\beta_{RE})]^{-1}(\hat\beta_{FE} - \hat\beta_{RE}) \xrightarrow{d} \chi^2_K$ tests whether FE and RE estimates differ systematically. Rejection means $\alpha_i$ is correlated with $X_{it}$, favoring FE.
:::

```{r}
#| label: hausman-test
phtest(fe_fit, re_fit)
```

A small p-value rejects the RE assumption---state fixed effects are correlated with the regressors, so FE is preferred.

**Caveat:** Using the Hausman test to *choose* between FE and RE creates a pretest estimator with distorted coverage. The research design should determine the estimator; report the Hausman test as a diagnostic. The Hausman test can also be viewed as a [GMM overidentification test](ch11-gmm.qmd#sec-j-test).

## Correlated random effects (Mundlak) {#sec-mundlak}

Mundlak (1978) proposed a middle ground: include the group means $\bar{x}_i$ as additional regressors in the RE model. This "soaks up" the correlation between $\alpha_i$ and $x_{it}$:

$$\alpha_i = \delta_0 + \bar{x}_i' \delta + \zeta_i, \quad \mathbb{E}[\zeta_i \mid x_{it}, z_i] = 0$$

```{r}
#| label: cre-mundlak
# Add group means of time-varying regressors
pdata$lpcap_bar <- Between(log(pdata$pcap))
pdata$lpc_bar   <- Between(log(pdata$pc))
pdata$lemp_bar  <- Between(log(pdata$emp))
pdata$unemp_bar <- Between(pdata$unemp)

# CRE = RE + group means
cre_fit <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp +
                 lpcap_bar + lpc_bar + lemp_bar + unemp_bar,
               data = pdata, model = "random")

# CRE slopes = FE slopes (by FWL)
cat("CRE slopes: ", round(coef(cre_fit)[2:5], 4), "\n")
cat("FE slopes:  ", round(coef(fe_fit), 4), "\n")
```

The CRE slopes on time-varying regressors match FE exactly. This follows from the [FWL theorem](ch04-sensitivity.qmd#thm-fwl): partialing out $\bar{x}_i$ from $x_{it}$ gives the within-demeaned data.

::: {#thm-mundlak}
## Correlated Random Effects (Mundlak)
Adding group means $\bar{x}_i$ to the RE model gives CRE: $y_{it} = x_{it}'\beta + \bar{x}_i'\delta + \alpha_i + \varepsilon_{it}$. The slope $\hat\beta_{CRE}$ equals $\hat\beta_{FE}$ by FWL. Testing $\delta = 0$ is equivalent to the Hausman test but works with robust SEs and allows estimation of time-invariant variable effects.
:::

::: {.callout-note}
## CRE = FE + Time-Invariant Variables
The Mundlak/CRE approach gives identical slope coefficients to FE while also allowing estimation of effects of time-invariant regressors (region, sex, ethnicity). It also provides a robust version of the Hausman test through an F-test on the group means.
:::

### The Mundlak test = Hausman test

Testing $H_0: \delta = 0$ (the group means have no additional explanatory power) is equivalent to the Hausman test:

```{r}
#| label: mundlak-test
# F-test on the group means
library(car)
linearHypothesis(cre_fit,
  c("lpcap_bar = 0", "lpc_bar = 0", "lemp_bar = 0", "unemp_bar = 0"))
```

### CRE identifies effects of time-invariant variables

A key advantage of CRE over FE: it can estimate effects of time-invariant regressors. FE sweeps them out along with $\alpha_i$.

```{r}
#| label: cre-time-invariant
# Add a time-invariant variable: region
pdata$region_ne <- as.numeric(pdata$region == "1")
pdata$region_s  <- as.numeric(pdata$region == "3")
pdata$region_w  <- as.numeric(pdata$region == "4")

cre_region <- plm(log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp +
                    lpcap_bar + lpc_bar + lemp_bar + unemp_bar +
                    region_ne + region_s + region_w,
                  data = pdata, model = "random")

# Region coefficients (relative to North Central)
cat("Region effects (CRE):\n")
cat("  Northeast:", round(coef(cre_region)["region_ne"], 4), "\n")
cat("  South:    ", round(coef(cre_region)["region_s"], 4), "\n")
cat("  West:     ", round(coef(cre_region)["region_w"], 4), "\n")
```

## The problem with TWFE under staggered treatment {#sec-twfe-bias}

Standard two-way fixed effects (TWFE) with staggered treatment can produce severely biased estimates when treatment effects are heterogeneous. TWFE implicitly compares newly-treated units to already-treated units, creating negative weights on some group-time ATTs.

::: {#thm-twfe-bias}
## TWFE Bias Under Staggered Treatment
With staggered adoption and heterogeneous treatment effects, TWFE assigns negative weights to some group-time ATTs — using already-treated units as controls for newly-treated units. The resulting estimate can be far from any meaningful causal parameter, including the wrong sign.
:::

::: {.callout-warning}
## TWFE Can Produce Negative Weights
Under staggered treatment with dynamic or heterogeneous effects, TWFE regression assigns negative weights to some group-time treatment effects. This means the TWFE coefficient is not a convex average of individual treatment effects — it can even have the wrong sign. Use Callaway-Sant'Anna, Sun-Abraham, or fect instead.
:::

### Simulation: TWFE bias

```{r}
#| label: twfe-bias-sim
set.seed(42)
N <- 200; T_max <- 10

# Staggered adoption: 3 cohorts + never-treated
cohorts <- c(4, 6, 8)  # treatment start times
sim_panel <- data.frame()

for (i in 1:N) {
  # Assign to cohort (25% each)
  g <- sample(c(cohorts, 0), 1)  # 0 = never treated
  alpha_i <- rnorm(1, sd = 2)

  for (t in 1:T_max) {
    treated <- (g > 0) & (t >= g)
    # Treatment effect grows with exposure AND varies by cohort
    te <- if (treated) 1.0 + 0.3 * (t - g) + 0.5 * (g == 4) else 0
    y <- 2 + alpha_i + 0.2 * t + te + rnorm(1)

    sim_panel <- rbind(sim_panel,
      data.frame(id = i, time = t, y = y,
                 treat = as.integer(treated),
                 cohort = g,
                 first_treat = ifelse(g == 0, 0, g)))
  }
}

# TWFE regression
twfe <- feols(y ~ treat | id + time, data = sim_panel)
cat("TWFE estimate:", round(coef(twfe)["treat"], 3), "\n")

# True average ATT among treated observations
true_att <- mean(sim_panel$y[sim_panel$treat == 1]) -
  mean(2 + ave(rnorm(N), rep(1:N, each = T_max)) + 0.2 * sim_panel$time[sim_panel$treat == 1])
cat("(The true ATT varies by cohort and exposure time -- TWFE averages with potentially negative weights)\n")
```

## Callaway--Sant'Anna: group-time ATTs

The `did` package (Callaway & Sant'Anna, 2021) avoids the TWFE problem by estimating separate ATTs for each group-time pair, then aggregating. The `mpdta` dataset tracks teen employment across 500 U.S. counties, with staggered minimum wage increases.

```{r}
#| label: mpdta-setup
data(mpdta)
cat("Counties:", length(unique(mpdta$countyreal)),
    " Years:", paste(range(mpdta$year), collapse = "-"),
    " Treatment cohorts:", paste(sort(unique(mpdta$first.treat[mpdta$first.treat > 0])),
                                  collapse = ", "), "\n")
head(mpdta)
```

### Estimating group-time ATTs

```{r}
#| label: cs-attgt
# Estimate all group-time ATTs
cs_out <- att_gt(
  yname = "lemp",
  gname = "first.treat",
  idname = "countyreal",
  tname = "year",
  xformla = ~1,              # unconditional
  data = mpdta,
  control_group = "nevertreated",
  est_method = "dr"          # doubly robust (default)
)

summary(cs_out)
```

Each row is an ATT(g, t): the average treatment effect for group $g$ (defined by when they were first treated) at time $t$.

```{r}
#| label: cs-attgt-plot
ggdid(cs_out, ylim = c(-0.3, 0.3))
```

### Aggregation: event study

The many group-time ATTs can be aggregated into an event study (dynamic effects by exposure time):

```{r}
#| label: cs-event-study
cs_dyn <- aggte(cs_out, type = "dynamic")
summary(cs_dyn)
```

```{r}
#| label: cs-event-plot
#| fig-cap: "Callaway-Sant'Anna event study: pre-treatment estimates near zero support parallel trends"
ggdid(cs_dyn, ylim = c(-0.3, 0.3))
```

Pre-treatment estimates near zero support the parallel trends assumption. Post-treatment estimates show the dynamic treatment effect.

### Simple aggregation: overall ATT

```{r}
#| label: cs-simple
cs_simple <- aggte(cs_out, type = "simple")
summary(cs_simple)
```

### Group-specific effects

```{r}
#| label: cs-group
cs_group <- aggte(cs_out, type = "group")
summary(cs_group)
```

Different cohorts may experience different effects---early adopters vs. late adopters of the minimum wage increase.

### With covariates

Parallel trends can be conditioned on covariates. Here we control for log population:

```{r}
#| label: cs-covariates
cs_cov <- att_gt(
  yname = "lemp",
  gname = "first.treat",
  idname = "countyreal",
  tname = "year",
  xformla = ~lpop,           # conditional on log population
  data = mpdta,
  control_group = "nevertreated"
)

cs_cov_dyn <- aggte(cs_cov, type = "dynamic")
ggdid(cs_cov_dyn, ylim = c(-0.3, 0.3))
```

## Counterfactual estimators with fect

The `fect` package (Liu, Wang & Xu, 2024) takes a different approach: it imputes counterfactual outcomes for treated observations by fitting a model on untreated observations only, then computes treatment effects as the difference between observed and imputed outcomes.

The `simdata` dataset has 200 units over 35 periods with treatment that can switch on and off:

```{r}
#| label: fect-data
data(fect)
cat("Units:", length(unique(simdata$id)),
    " Periods:", paste(range(simdata$time), collapse = "-"),
    " Treated obs:", sum(simdata$D), "\n")
```

### Two-way FE counterfactual

The simplest `fect` estimator uses two-way fixed effects to impute counterfactuals:

```{r}
#| label: fect-fe
out_fe <- fect(Y ~ D + X1 + X2, data = simdata,
               index = c("id", "time"),
               method = "fe", force = "two-way",
               se = TRUE, nboots = 200,
               parallel = FALSE)
```

```{r}
#| label: fect-fe-summary
cat("ATT estimate:", round(out_fe$est.avg[1, 1], 3), "\n")
cat("SE:          ", round(out_fe$est.avg[1, 2], 3), "\n")
cat("95% CI:      [", round(out_fe$est.avg[1, 3], 3), ",",
    round(out_fe$est.avg[1, 4], 3), "]\n")
```

### Event study (gap) plot

The default plot shows period-by-period ATTs relative to treatment onset:

```{r}
#| label: fect-fe-plot
plot(out_fe, main = "FE counterfactual: ATT by periods since treatment")
```

### Interactive fixed effects (IFE)

When parallel trends may not hold, interactive fixed effects allow for unit-specific responses to common latent factors:

$$Y_{it}(0) = \alpha_i + \xi_t + f_t' \lambda_i + X_{it}'\beta + \varepsilon_{it}$$

The number of factors $r$ is selected by cross-validation:

```{r}
#| label: fect-ife
out_ife <- fect(Y ~ D + X1 + X2, data = simdata,
                index = c("id", "time"),
                method = "ife", force = "two-way",
                CV = TRUE, r = c(0, 5),
                se = TRUE, nboots = 200,
                parallel = FALSE)
```

```{r}
#| label: fect-ife-summary
cat("IFE ATT:", round(out_ife$est.avg[1, 1], 3),
    " (selected r =", out_ife$r.cv, ")\n")
```

```{r}
#| label: fect-ife-plot
plot(out_ife, main = "IFE counterfactual: ATT by periods since treatment")
```

### Matrix completion

Matrix completion treats the $N \times T$ matrix of untreated potential outcomes as approximately low-rank and uses nuclear norm regularization to impute missing entries---no need to specify the number of factors:

```{r}
#| label: fect-mc
out_mc <- fect(Y ~ D + X1 + X2, data = simdata,
               index = c("id", "time"),
               method = "mc", force = "two-way",
               CV = TRUE,
               se = TRUE, nboots = 200,
               parallel = FALSE)
```

```{r}
#| label: fect-mc-summary
cat("MC ATT:", round(out_mc$est.avg[1, 1], 3), "\n")
```

### Comparing methods

```{r}
#| label: fect-comparison
methods_comp <- data.frame(
  Method = c("FE", "IFE", "MC"),
  ATT = round(c(out_fe$est.avg[1, 1], out_ife$est.avg[1, 1], out_mc$est.avg[1, 1]), 3),
  SE = round(c(out_fe$est.avg[1, 2], out_ife$est.avg[1, 2], out_mc$est.avg[1, 2]), 3),
  CI_lower = round(c(out_fe$est.avg[1, 3], out_ife$est.avg[1, 3], out_mc$est.avg[1, 3]), 3),
  CI_upper = round(c(out_fe$est.avg[1, 4], out_ife$est.avg[1, 4], out_mc$est.avg[1, 4]), 3)
)
methods_comp
```

### Placebo test for pre-trends

The placebo test checks whether there are "treatment effects" in the pre-treatment period---evidence against parallel trends:

```{r}
#| label: fect-placebo
out_placebo <- fect(Y ~ D + X1 + X2, data = simdata,
                    index = c("id", "time"),
                    method = "fe", force = "two-way",
                    placeboTest = TRUE,
                    placebo.period = c(-2, 0),
                    se = TRUE, nboots = 200,
                    parallel = FALSE)
plot(out_placebo, main = "Placebo test: pre-treatment effects should be zero")
```

### Equivalence test

The equivalence test provides a more formal assessment: are pre-treatment ATTs close enough to zero?

```{r}
#| label: fect-equiv
plot(out_fe, type = "equiv",
     main = "Equivalence test for pre-treatment periods")
```

## Matrix completion for panel data (conceptual)

Athey, Bayati, Doudchenko, Imbens & Khosravi (2021) formalize the connection between DID, synthetic control, and matrix completion. The key insight is that all three are special cases of the same framework, differing in what structure they impose on the matrix of untreated potential outcomes $Y(0)$:

| Method | Assumption on $Y(0)$ | Best when |
|--------|---------------------|-----------|
| DID | Additive: $Y_{it}(0) = \alpha_i + \xi_t$ (rank 1) | Strong parallel trends |
| Synthetic control | Hard rank constraint on columns | Few treated units, many controls |
| Matrix completion | Soft low-rank via nuclear norm penalty | General (robust across settings) |

The `MCPanel` package (available from GitHub: `susanathey/MCPanel`) implements nuclear norm minimized matrix completion. The idea is to find a low-rank matrix $L$ plus unit and time effects that best approximate the observed (untreated) entries:

$$\min_{L, \gamma, \delta} \frac{1}{|\mathcal{O}|} \sum_{(i,t) \in \mathcal{O}} (Y_{it} - L_{it} - \gamma_i - \delta_t)^2 + \lambda \|L\|_*$$

where $\|L\|_*$ is the nuclear norm (sum of singular values) and $\lambda$ is chosen by cross-validation. When $\lambda$ is very large, $L \to 0$ and the method reduces to DID. When $\lambda$ is small, it approximates interactive fixed effects.

We can demonstrate the idea with a simple simulation:

```{r}
#| label: mc-concept
set.seed(99)
N_mc <- 30; T_mc <- 20; R_true <- 2

# Low-rank structure
A <- matrix(rnorm(N_mc * R_true), N_mc, R_true)
B <- matrix(rnorm(T_mc * R_true), T_mc, R_true)
L_true <- A %*% t(B)

# Unit and time effects
gamma <- rnorm(N_mc, sd = 0.5)
delta <- rnorm(T_mc, sd = 0.5)
Y0 <- L_true + outer(gamma, rep(1, T_mc)) + outer(rep(1, N_mc), delta) +
  matrix(rnorm(N_mc * T_mc, sd = 0.3), N_mc, T_mc)

# Treatment: units 1-10 treated from period 15 onward
tau_true <- 2.0  # constant treatment effect
Y_obs <- Y0
Y_obs[1:10, 15:T_mc] <- Y_obs[1:10, 15:T_mc] + tau_true

# What DID would estimate
did_control_pre <- mean(Y_obs[11:N_mc, 1:14])
did_control_post <- mean(Y_obs[11:N_mc, 15:T_mc])
did_treat_pre <- mean(Y_obs[1:10, 1:14])
did_treat_post <- mean(Y_obs[1:10, 15:T_mc])
did_att <- (did_treat_post - did_treat_pre) - (did_control_post - did_control_pre)

# SVD-based imputation (simplified matrix completion)
# Use control rows + pre-treatment treated rows to impute
mask <- matrix(1, N_mc, T_mc)
mask[1:10, 15:T_mc] <- 0  # missing = treated

# Impute using low-rank SVD of observed entries
Y_masked <- Y_obs * mask
Y_masked[1:10, 15:T_mc] <- NA

# Simple imputation: use SVD on complete rows to predict treated post
Y_control <- Y_obs[11:N_mc, ]
svd_control <- svd(Y_control)
# Project treated pre-period onto control space
Y_treat_pre <- Y_obs[1:10, 1:14]
# Use first R_true factors
V_pre <- svd_control$v[1:14, 1:R_true]
V_post <- svd_control$v[15:T_mc, 1:R_true]
U_hat <- Y_treat_pre %*% V_pre %*% solve(t(V_pre) %*% V_pre)
Y0_hat_post <- U_hat %*% t(V_post)

# Adjust for level (add mean)
mc_att <- mean(Y_obs[1:10, 15:T_mc] - Y0_hat_post) -
  mean(Y_obs[1:10, 1:14] - U_hat %*% t(V_pre))

cat("True ATT:     ", tau_true, "\n")
cat("DID estimate: ", round(did_att, 3), "\n")
cat("MC estimate:  ", round(mc_att, 3), "\n")
```

When the data have a low-rank structure that violates simple parallel trends, matrix completion can outperform DID by exploiting the factor structure.

## When to use which method

The landscape of DiD and panel methods has expanded. Here is a practical guide:

```{r}
#| label: method-guide
#| echo: false
cat("
Modern DiD / Panel Method Guide:
==================================

STANDARD PANEL (no staggered treatment):
  - alpha_i correlated with X?
    Yes --> Fixed Effects or CRE (Mundlak)
    No  --> Random Effects (rare in observational data)
  - Lagged dependent variable?
    Yes --> Arellano-Bond or System GMM (see Ch. 12)

STAGGERED DiD (treatment adopted at different times):
  - Homogeneous treatment effects?
    Yes --> Standard TWFE is fine
    No  --> Use modern methods:
      * Callaway-Sant'Anna (did package):
        - Group-time ATTs, flexible aggregation
        - Supports covariates, doubly robust estimation
        - Requires staggered adoption (no treatment reversal)
      * fect package:
        - FE, IFE, or matrix completion counterfactuals
        - Handles treatment reversal
        - Built-in placebo and equivalence tests
        - Also wraps other methods (CS, Sun-Abraham, etc.)
      * MCPanel:
        - Matrix completion (nuclear norm)
        - Unifies DID, synthetic control, and IFE
        - Best when outcome matrix is approximately low-rank

DIAGNOSTICS:
  - Always check pre-trends (event study / placebo test)
  - For GMM: Sargan/J-test + AR(2) test
  - For RE: Hausman test (as diagnostic, not decision rule)
  - For CRE: F-test on group means = robust Hausman test
")
```

## Summary

- **Random effects** assumes $\alpha_i \perp X_{it}$ and applies GLS with the composite error structure. It is more efficient than FE when valid, but inconsistent when violated.
- **The Hausman test** compares FE and RE; use it as a diagnostic, not a model-selection tool.
- **Correlated random effects (Mundlak)** nests both FE and RE: it recovers FE slopes while also estimating effects of time-invariant variables. The Mundlak test ($\delta = 0$) equals the Hausman test but works with robust SEs.
- **TWFE fails under staggered treatment** with heterogeneous effects due to negative weighting.
- **Callaway--Sant'Anna** (`did` package) estimates group-time ATTs and aggregates them into event studies, group-specific effects, or an overall ATT---avoiding the negative weighting problem.
- **fect** provides counterfactual imputation estimators: FE, interactive fixed effects, and matrix completion. It handles treatment reversal and includes built-in diagnostics.
- **Matrix completion** (Athey et al.) unifies DID and synthetic control as special cases of nuclear norm minimization, providing a flexible middle ground.
- All panel estimators connect to the **GMM framework**: FE uses within-group moments, RE adds between-group moments, CRE augments RE to nest FE, and dynamic panels use lagged instruments.
